---
output: 
  stevetemplates::article:
    fig_caption: true
bibliography: ref.bib
biblio-style: apsr
title: "Elusive Collaboration? The Determinants of Lead Donorship in International Development"
#thanks: "**Corresponding author**: williamsmd@denison.edu. The author has no conflict of interest to report and did not receive funding for this research. Data and replication materials are available on the author's GitHub page at: https://github.com/milesdwilliams15. In addition to the three annonymous reviewers, I would like to thank Matthew S. Winters, Xinyuan Dai, Robert D. Pahre, Robert J. Carroll, Alice Iannantuoni, Tobias Heinrich, and many others who provided feedback on previous and related iterations of this manuscript."
# author:
#  - name: Miles D. Williams
#  - affiliation: Data for Political Research, Denison University, Granville, OH, USA.
abstract: "**Abstract**: Lead donorship is a condition of stustained outsided responsibility by one foreign aid donor country for giving aid to a developing country. Prior research argues that lead donorship can either be a product of donors collaborating to carve out spheres of influence to maximize non-development foreign policy goals or else of donors buck-passing responsibility for addressing recipient needs. Which is it? This study seeks to answer this question using theory-driven expectations about when and where lead donorship should arise if donors cooperate or systematically fail to do so. It further develops novel measures of foreign aid's public goods characteristics and donors' marginal returns from giving aid that should help triangulate conditions under which we can expect lead donorship to arise due to cooperation or else its absence. Estimates from a flexible semi-parametric modeling strategy show that when and where lead donorship has the highest probability of occurring is consistent with systematic cooperation failure in the allocation of foreign aid. Various robustness checks are done and alternative explanations are also discussed. These findings provide insight into ongoing policy debates about donor cooperation and aid effectiveness and raise questions about what deeper strategic logics preclude donor cooperation."
keywords: foreign aid, lead donorship, cooperation
# date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: cochineal
fontsize: 12pt
spacing: double
endnote: no
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE,
                      echo=FALSE,
                      message=FALSE, warning=FALSE,
                      fig.path='figs/',
                      cache.path = '_cache/',
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })
library(tidyverse)
library(coolorrr)
set_theme()
set_palette()
library(estimatr)
library(texreg)
```

\clearpage

# Introduction

Why does lead donorship—a condition of sustained outsized responsibility
by one bilateral foreign aid donor government for providing development
assistance to a developing country—arise when and where it does? Does it
emerge because of donor collaboration or is it a reflection of an
uncooperative general equilibrium? Prior work suggests lead donorship
portends good news about donor cooperation but bemoans its relative
decline over the past few decades [@steinwand2015]. However, for good
news about lead donorship to be good news about cooperation, we should
expect to see a distinct pattern in when and where lead donorship
arises. This study addresses this issue by modeling patterns in lead
donorship as a function of novel measures of recipient need and
donor-recipient ties to help triangulate different foreign policy goals
of donors in aid recipients. Following the theoretical framework
outlined by @steinwand2015, two competing hypotheses (*collaboration*
and *elusive collaboration*) for how donor-recipient ties and recipient
need interact to predict patterns in lead donorship are pitted against
each other. One hypothesis specifies what the distribution of lead
donorship should look like if donors systematically cooperate in the
allocation of bilateral aid while the other specifies what the
distribution of lead donorship should look like if donors systematically
fail to cooperate. In the main analysis, the *elusive collaboration*
hypothesis finds support over the *collaboration* hypothesis, and
further robustness checks support this conclusion. In short, the pattern
of lead donorship is consistent with systematic cooperation failure in
the allocation of foreign aid. This novel empirical finding is
significant for both normative and theoretical reasons.

First, identifying whether lead donorship is a product of cooperation or
cooperation failure is important because of its connection to aid
fragmentation, which characterizes the extent to which a developing
country receives aid from a diverse portfolio of donors. Past findings
are mixed, but many propose that aid fragmentation may strain recipient
bureaucratic capacity [@kanbur2006; @knackRahman2007], increase donor
transaction costs [@anderson2012], and reduce a donor's ability to use
aid conditionality to spur positive development and governance reforms
in recipients [@hernandez2017; @zeitz2021]. At the same time, others
propose that lack of recipient choice (i.e., low fragmentation or the
presence of a lead donor) can have negative effects, and that under the
right conditions, fragmentation actually has pro-development and growth
impacts [@gehringEtAl2017; @gutting2017donor]. While a definitive
statement on the effect of fragmentation remains elusive, most scholars
agree that it matters for aid effectiveness. By extension so does lead
donorship, because as @steinwand2015 argues, why lead donorship emerges
when and where it does may play a role in conditioning the impact of aid
fragmentation on aid effectiveness.

Second, a question of major theoretical importance is understanding
whether and how cooperation in the allocation of foreign aid is
possible. Many studies that consider this issue take a spatial modeling
approach, using the correlation in the giving of a donor with that of
its peers to infer cooperation or cooperation failure [for example,
@fuchsEtAl2015; @steinwand2011; @steinwand2015]. Such an approach has
many advantages but may be sensitive to biases induced by endogeneity,
circular causation, and heterogeneity in the spatial correlation of
donor giving. The latter concern is especially significant given the
diverse objectives donor governments pursue through their aid
allocation. Using lead donorship to study this issue provides a fresh
approach that side-steps the need to estimate the spatial correlation in
donor giving to infer donor behavior. Instead, it offers a chance to
consider the *general equilibrium* implications of donor behavior
whereas a classical spatial modeling approach provides only a *partial*
picture.

The paper proceeds as follows. First, the concept of lead donorship is
introduced. Credit for the concept is owed to @steinwand2015. Though
important, lead donorship has received little engagement in subsequent
research. This is unfortunate, because the concept has untapped
potential for identifying whether systematic cooperation or cooperation
failure exists in the aid allocation decisions of donor governments.

Next is a theoretical discussion of what patterns in lead donorship we
should observe depending on whether donors cooperate or fail to do so in
the allocation of foreign aid. This discussion addresses two buckets of
goods donors seek from their aid allocation: *public* and *private.*
When and where aid produces public goods for donor governments, classic
theories of collective action [@olson2012] imply that cooperation should
reduce the likelihood of lead donorship. Meanwhile, cooperation failure
should increase the likelihood of lead donorship in proportion to the
degree to which marginal returns from aid are concentrated in a single
donor [@buchholz2016olson]. The opposite is true when and where aid
produces rival goods. Competition for rival benefits, like political
influence or market access, should lead to crowding in, making it more
difficult for the donor with the greatest marginal returns to maintain
the top position. Cooperation implies, to the contrary, that donors
agree to forego competition and thus cede the top position to the donor
with the most to gain from its relationship with a developing country.

These two competing behavioral hypotheses---which I call *collaboration*
and *elusive collaboration*, respectively---are pitted against each
other in the remainder of the paper. The next sections discuss these
hypotheses in more detail and empirical hurdles that need to be
surmounted in order to test them. The main hurdle is how to capture when
one kind of good (public or private) is most pronounced and whether
marginal returns from the good are concentrated in a single donor or
enjoyed by many. Two concepts are introduced: *donor interest* or *ties*
with developing countries and *recipient need*. I argue that these two
concepts permit triangulating when and where the public goods properties
of aid will be more pronounced (high need) and when and where marginal
returns are concentrated in a single donor (high concentration of dyadic
donor-recipient ties).

The study then turns to measurement. First, lead donorship is
operationalized and some possible limitations discussed. Next,
donor-recipient ties and recipient need are considered. Most studies
measure these concepts using vectors of multiple different variables.
The approach taken here is to combine each into a unique measure. I rely
on two sets of variables to construct an index of donor ties with
recipients and an index of recipient need, respectively. The first is
composed of measures of bilateral trade, colonial past, bilateral
distance, and military alliances. The second is composed of recipient
level measures of average income, population size, ongoing civil war,
the impact of natural disasters, and strength of democracy. A method to
combine these variables into their respective measures is introduced
that creates a linear combination that is designed to maximally covary
with its components.

Buttressing these measures are two assumptions. The first is that the
degree of recipient need is proportional to the extent to which
addressing recipient problems with foreign aid addresses a public goods
problem (e.g., promoting economic growth, limiting drivers of global
migration, and supporting political stability). The second assumption is
that the extent to which a donor has relatively stronger ties to a
recipient is proportional to its marginal returns for giving aid to a
recipient (e.g., close trade ties, geographic proximity, colonial
legacy, and so on). These general buckets of donor considerations in
giving aid to developing countries are well-studied. However, where need
and ties were often pitted against each other in the empirical
literature, more recent work shows that these drivers of aid giving are
complementary [@bermeo2017; @bermeo2018]. For example, strength of donor
ties with a recipient can mediate the benefits donors receive from
promoting public goods in a recipient with foreign aid. Conversely, when
recipient need is low, donor ties can determine the relative benefits
donors receive from using aid for non-development purposes.

Based on the theoretical discussion, we should observe one of two
patterns in lead donorship as a function of bilateral ties and recipient
need. If donors cooperate, lead donorship should occur with the greatest
frequency in recipients where we observe low need and ties concentrated
with a single donor. Conversely, if donors fail to cooperate, lead
donorship should occur with the greatest frequency in recipients where
we observe high need and ties concentrated with a single donor.

These predictions imply an interactive hypothesis between need and ties,
but one that is inconsistent with a simple two-way interaction in a
linear model. Therefore, a semi-parametric approach is adopted. The data
are aggregated to the recipient-year level with lead donorship modeled
within a generalized additive model or GAM framework using a logit link
function with year fixed effects and random recipient effects. A joint
smoothing function is used to map the interaction of a measure of donor
tie concentration and level of recipient need to the probability of lead
donorship in a developing country.

The results show that lead donorship has the greatest probability of
occurrence in developing countries that have high need and that have
bilateral ties concentrated with one donor government. This finding is
consistent with the *elusive collaboration* hypothesis, meaning that the
pattern in lead donorship is consistent with the presence of
uncooperative competition for top donor status where private goods are
predominant and buck-passing where public goods are predominant.

It is important to note that while this finding is consistent with the
theoretical argument, other explanations cannot be ruled out. Several
are discussed following the presentation of the main results, including
the possibility that lead donorship is the product of coordination or
that it is epiphenominal to donor interests. Robustness checks are done
as well that consider alternative measurement strategies, a lagged
version of donor-recipient ties to account for possible endogeneity, and
a more restrictive measure of lead donorship. Some exceptional cases of
lead donorship are discussed as well. These are instances of lead
donorship that arise when and where the theoretical argument contends it
should be absent. Rather than suggest limitations with the theoretical
argument, these cases appear to show that some cases are not adequately
captured by the measure of donor-recipient ties. In short, these
exceptions prove the rule, revealing instead that the measures developed
in this study are somewhat blunt (an admitted limitation with any
measurement approach that collapses many factors into one).

Ultimately, these findings matter for ongoing policy debates and raise
important theoretical questions. From a policy perspective, the seeming
failure to identify much collaborative behavior among donors (at least
at a macro level) is frustrating in light of decades-long efforts to
spur greater donor cooperation. From a theoretical perspective, systemic
cooperation failure begs the question as to why uncooperative behavior
is so endemic. Are there features of the strategic environment that
donors and recipients inhabit that prevent cooperation? In addition, is
aid effectiveness improved or worsened by virtue of when and where lead
donorship emerges? These are important questions for future research to
address.

# Lead Donorship and Its Connection to Donor Cooperation

Lead donorship characterizes a scenario where a single donor has a
long-term relationship with an aid recipient and is responsible for the
greatest share of foreign aid given [@steinwand2015]. The concept is
closely connected to aid fragmentation, which is a measure of the extent
to which a developing country receives aid from a diverse portfolio of
donors. However, fragmentation is not strictly the inverse of lead
donorship. While low levels of fragmentation are an important condition
for lead donorship to emerge, low fragmentation is just one dimension of
lead donorship. Lead donorship is not just a reflection of a
concentrated aid donor portfolio for a recipient, it is has a *temporal*
dimension as well. As @steinwand2015 notes, "lead donorship [is] a
long-time stable exclusive relationship between one donor country and a
recipient country, in which a donor continuously acts as biggest
provider of foreign aid" (445). Based on this definition, while high
fragmentation would mean the absence of a lead donor, low fragmentation
is not sufficient to say a lead donor exists.

One of the main questions surrounding lead donorship is whether it is a
product of donor collaboration in carving out spheres of influence, or
else if it is a product of uncooperative buck-passing. The key factor
contextualizing lead donorship is the public or rival nature of the
goals donor governments use foreign aid to achieve. As @dreher2024aid
note in a recent overview of the aid literature, industrialized
countries use foreign aid as a multi-purpose foreign policy tool that
can serve both as an instrument for promoting economic development and
for providing donor governments with political influence, market access,
and security. The former is often thought to have public goods
properties, while the latter group of objectives is thought to have
private goods properties.

When dealing with public goods, theories of collective action hold that
if actors fail to cooperate, the actor with the greatest marginal
returns from a public good will end up bearing the greatest share of the
burden in supplying it [@buchholz2016olson; @olson2012]. By their very
nature, public goods are nonrival and nonexcludable. That means one
actor's enjoyment of the good does not come at the expense of another's,
and no actor can be excluded from enjoying the good, even if they do not
directly pay for it. This creates adverse incentives for all involved.
Not only does each actor (regardless of its marginal returns) generally
pay less for the public good than they would absent free-riding, but
also the actors with lower marginal returns from the good will pay
disproportionately less than the actor with the greatest marginal
returns.

The welfare implications of free-riding are simple: all actors would be
better off if they did not free-ride. As @steinwand2015 notes, in the
context of foreign aid, "coordination in the provision of aid with
public goods properties helps to reduce free-riding and increases aid
levels, with ultimately positive results for development" (444-445).
That means that cooperation actually increases aid giving by all donors
and thus lowers the likelihood of lead donorship. In short, with respect
to addressing public goods, aid effectiveness is improved by limiting
lead donorship, but donor self-interest makes lead donorship more
likely.

The converse holds for rival goods. Here, lack of cooperation manifests
as a race to the top that makes it harder for the actor with the
greatest marginal returns from rival goods to maximize those returns.
The result is greater aid fragmentation and thus a lower likelihood of
lead donorship. Like free-riding, competition is inefficient from a
donor welfare perspective because it leads donors to crowd in resources
that could otherwise be used to achieve other goals. It may also be bad
from a recipient development perspective because by expanding recipient
choice, rents shift away from the donor in favor of the recipient
government. But, as @steinwand2015 notes, if donors cooperate to limit
competition, "[d]onor fragmentation decreases, lead donorship arises,
but overall aid amounts fall" with the end result being less waste and
overlap in aid projects and less opportunity for rent-seeking by
recipient governments (445). In short, with respect to rival goods, aid
effectiveness is improved with lead donorship, but donor self-interest
limits the possibility of a lead donor emerging.

@buchholz2016olson present a formal model of global public goods that
may help to clarify the logic at work here and further show when and why
the good in question will either increase or decrease the likelihood
that a lead donor emerges if donors fail to collaborate.
@buchholz2016olson use a very simple game theoretic set-up that allows
for a pair of actors in the provision of a public good to differ with
respect to three things: their budget, their marginal returns from the
public good (preferences), and their comparative advantage (or ability)
in supplying the public good. @buchholz2016olson find that if you take
two actors involved in the supply of a public good who have similar
budgets but asymmetric preferences or comparative advantages, the actor
with the stronger preference or comparative advantage will shoulder most
or all of the burden of supplying the good while the other actor
free-rides. As @buchholz2016olson further find, the influence of
preference and comparative advantage can be enough such that an actor
with a comparatively smaller budget will nonetheless bear the greatest
responsibility for supplying the public good if they are more motivated
(stronger preference) or better equipped (comparative advantage) to do
so. However, when actors are similar with respect to budget, preference,
and comparative advantage, we should expect overall supply of the good
to be lower, but no one actor to bear greater responsibility for
supplying it. In short, there is nothing to compel one actor to take the
lead thus making it possible for the other to sit on the sidelines.

While rival goods are not addressed by @buchholz2016olson, by reversing
the sign of the impact that actor spending has on the other actor's
ability to enjoy the good in question, solving for actors'
best-responses shows that they now would be motivated to give more when
the other does as well. Furthermore, where asymmetrical preferences or
comparative advantages create space for a single actor to emerge as the
chief supplier of a public good, these same factors will only enhance
competition. This is because, with asymmetrical preferences, the actor
with the weaker preference for the rival good faces strong incentives to
keep up with the other actor. The mechanism driving this is that the
actor with stronger preferences for a rival good will naturally spend
more on that good to begin with. This forces the other actor to spend
more on the rival good than otherwise in order to ensure they, too, get
to enjoy some of the benefits they desire from the rival good. The
result is crowding-in, with the actor that has the strongest preference
for the rival good spending a little more than it would otherwise and
the actor that has the weakest preference for the rival good spending a
lot more than it would otherwise. In this way, competition narrows the
gap in spending by actors for the rival good. The spending gap will
similarly be slim or non-existent when actors are similar in terms of
budget, preference, and comparative advantage because they will still
need to compete to maximize their returns from the rival good.

We can apply the above general logic to many scenarios, including to
foreign aid allocation and lead donorship. Simply treat the actors in
this framework as aid donors and the public or rival goods they seek to
promote as either development or non-development goals that donors have
for giving aid to a particular recipient. In the case where donors seek
to address recipient need (a goal with public goods properties), there
will be incentives to free-ride. Free-riding, furthermore, creates space
for a lead donor to emerge, particularly when a single donor
disproportionately benefits from promoting development and/or has a
comparative advantage in doing so. Conversely, when no donor draws
comparatively greater benefits or has a comparative advantage in
promoting development and/or when donors seek to maximize
non-development foreign policy objectives (goals with rival goods
properties), a lead donor is unlikely to emerge. This is either because
no donor is comparatively more motivated/able to supply public goods to
sufficiently crowd-out other donors, or because competition forces
donors to crowd-in aid to maximize their non-development foreign policy
interests regardless of which donor has the strongest preference for or
comparative advantage in promoting said foreign policy interests.

It may be helpful to consider a hypothetical example with two donors (A
and B) and three aid recipients (X, Y, and Z) where donors seek to
promote public goods by giving foreign aid. Say A draws greater returns
from giving aid to X, B draws greater returns from giving aid to Y, and
both draw similar returns from giving aid to to Z. The logic of public
goods laid out by @buchholz2016olson is consistent with an equilibrium
where A will specialize in giving aid to X, B will specialize in giving
aid to Y, and both will bear similar responsibility for giving aid to Z.
That is, A will be a lead donor in X, B will be a lead donor in Y, but
there will be no lead donor in Z.

Alternatively, suppose A and B seek to promote rival foreign policy
objectives in X, Y, and Z with their foreign aid. Assuming A still draws
greater returns from giving aid to X, B draws greater returns from
giving aid to Y, and both draw similar returns from giving aid to Z, we
will observe a different equilibrium emerge than with public goods.
Because of competition, even though B derives fewer benefits from rival
goods in X than does A, B will be compelled to give more aid to X than
it would otherwise, narrowing the gap between its giving and A's. The
reverse problem will happen in recipient Y. Meanwhile, both donors will
give equal amounts of aid to recipient Z. The end result is that no lead
donor is able to emerge across any of the recipient countries.

Steinwand's [-@steinwand2015] argument builds on the above logic and
extends it to consider what cooperation among donors would imply—that
is, if donors could stay their incentives to either free-ride or
compete. If cooperation failure implies lead donorship will arise when
and where addressing public goods is salient and when and where a single
donor has the greatest motivation and ability to do so, successful
cooperation implies that lead donorship will arise when and where the
pursuit of rival goods is salient and when and where a single donor has
the greatest motivation and ability to realize these rival goods.

The logic of lead donorship and its connection to cooperation is clear
enough once explained, but empirical identification of why lead
donorship emerges is quite the challenge. @steinwand2015, who was the
first to consider the concept of lead donorship, takes a novel and
creative approach. First, to delineate aid given primarily for public or
private goods, @steinwand2015 disaggregates foreign aid by two main
delivery channels: government-to-government and bypass aid.

@dietrich2013 is among the first to note the possibility of different
motives associated with these alternative forms of aid delivery.
@dietrich2013 argues that while donor governments certainly are driven
by non-development foreign policy goals, they also care about responding
to recipient country development need. She hypothesizes that donors
should therefore be sensitive to recipient governance quality,
calibrating their form of aid delivery to optimize the chances that aid
reaches the populations in a recipient country it is meant to help.
Specifically, she argues that donor countries will use the so-called
"bypass" channel in inverse proportion to the quality of governance in a
recipient.

Bypass aid encompasses resources that are given to nongovernmental
actors such as NGOs operating in recipients rather than through direct
cooperation with the state bureaucracy. In poorly governed developing
countries, donors may fear that aid will be lost to rent-seeking elites.
This mistrust in turn leads them to target more aid through the bypass
channel relative to the government-to-government channel. @dietrich2013
finds strong support for her claims, and later studies have found
similar patterns when examining different aid sectors [@bermeo2017;
@bermeo2018] and alternative measures of limited state capacity such as
civil conflict [@everettTirone2021]. Of course, later work by
@dietrich2016 puts important scope conditions on this practice, finding
that the political economies of donors---whether they are oriented
toward market-efficiency or a strong state---condition the use of the
bypass channel as well.

After separating aid by delivery channel, @steinwand2015 further applies
spatial autoregressive models and proposes alternative hypotheses for
the direction of the spatial correlation in donor giving on the basis of
delivery channel and the presence of donor collaboration. Specifically,
if donors cooperate in the provision of public goods, donor reactivity
to other-donor aid should be null when examining the bypass aid channel.
However, if donors do not collude, then they should give less aid in
inverse proportion to the giving of others due to free-riding.

Conversely, if donors collude in private goods aid, donors should have a
negative reaction to other-donor giving through the
government-to-government channel. The idea is that donors agree to a
delineation of spheres of influence. By foregoing competition for rival
goods where they reap only minimal returns, donors are free to target
aid in developing countries where they reap higher marginal returns.
But, if donors do not cooperate, they should crowd in resources in the
government channel and thus give more aid in proportion to the giving of
other donors.

Steinwand's findings only partially support these expectations. Most
notably, the fine-grained distinction between bypass and
government-to-government aid---that one is oriented toward public goods
and the other private---does not hold up in the empirical analysis. In
fact, @steinwand2015 finds evidence of competition in bypass aid,
contrary to the claim that the bypass channel closely corresponds to
public goods.

However, @steinwand2015 does claim to recover evidence that lead
donorship is linked to donor collaboration. He finds that with lead
donorship, a donor decreases its level of giving in proportion to the
giving of others. Conversely, donors increase their level of giving in
proportion to others in the absence of a lead donor. Importantly, at
least from Steinwand's perspective, this finding comes from examining
the government-to-government aid channel.

The conclusion drawn from this analysis is that competition is the
driving strategic concern in foreign aid allocation, but that lead
donorship suggests donors have successfully engaged in some amount of
cooperation. But @steinwand2015 adds to this the observation that lead
donorship has been on a secular decline over the past few decades, and
he thus calls for greater efforts to promote lead donorship to reduce
competitive waste.

Without calling into question the soundness of Steinwand's empirical
strategy, some caution in interpreting his results is warranted. In
particular, the observation that with lead donorship the spatial
correlation in donor giving is negative does not necessarily support the
argument that lead donorship arises out of donor cooperation. Steinwand
notes that bypass aid is surprisingly tied to competition for rival
goods. Is it not possible that government-to-government aid is partially
linked to public goods? How do we know that the pattern Steinwand labels
cooperation is not actually suggestive of free-riding?

The limitation of Steinwand's approach is that it relies heavily on aid
delivery channels to help contextualize donor motives and spatial
correlations in donor giving, which can be subject to heterogeneity,
endogeneity, and circular causation. If aid giving in general
(regardless of delivery method) reflects a mix of both public and
private goods, it is impossible to diagnose the meaning of spatial
correlations in donor giving by the presence or absence of lead
donorship. It may be that with lead donorship we observe a negative
correlation between individual and other-donor giving due to
collaboration. It also may be that when and where lead donorship arises,
it does because of free-riding. The observation of a negative spatial
correlation in donor giving in the presence of a lead donor is
consistent with both explanations, and lead donorship itself is likely a
descriptive product of the spatial correlation in donor giving.

In this study, an alternative empirical strategy is proposed. Rather
than use variation in lead donorship to explain donor reactions to the
giving of one another, it would be more useful to explain variation in
lead donorship by way of measures that help triangulate the publicness
or rivalness of the goods donor governments get out of the their aid
allocation. Based on the logic outlined earlier in this section, if lead
donorship emerges primarily when and where the public goods properties
of aid are greatest and where marginal returns are concentrated in but
one or a few donors, this would support the claim that lead donorship is
the product of uncooperative buck-passing. Alternatively, if lead
donorship emerges primarily when and where the rival goods properties of
aid are greatest and where marginal returns are concentrated in one or a
few donors, this would support the claim that lead donorship is the
product of donor collaboration. The key question, of course, is how to
identify which kind of good is at work and how best to capture the
returns different donors get from promoting these goods in recipients.
This is the subject of the next section.

# Triangulating Public and Rival Goods in Aid Allocation

The motivations behind giving foreign aid have intrigued IR scholars and
economists alike ever since the practice emerged in its modern form
after World War II. Early contributions to the literature were divided
with respect to whether the factors driving aid giving were primarily
humanitarian in nature, or else cynical, driven by the self-serving
foreign policy, economic, and security goals of donors.
@mckinlayLittle1979 set the tone for future studies by pitting measures
of recipient need against measures of donor interest in their analysis
of US aid data from 1960 to 1970. Future studies would adopt a similar
approach [see @alesinaDollar2000; @maizelsNissanke1984;
@schraederEtAl1998].

Contemporary research has since moved away from the sharp dichotomy
between recipient need and donor interest. Instead, scholars recognize
that donor governments' motives are often mixed and vary over time.
Existing studies point to a number of factors that correspond either
with donor interest or recipient need as drivers of donor giving.
Studies show that foreign aid is linked to donors' geostrategic goals,
interest in promoting bilateral trade, combating global terrorism,
maintaining influence over former colonies, gaining international
prestige, complementing military deployments, and addressing the root
causes of discontent and instability, to name but a few
[@alesinaDollar2000; @bearceTirone2010; @bermeo2017; @bermeo2018;
@kilbyDreher2010; @kisanganiPickering2015; @roundOdedokun2004;
@vanDerVeen2011].

Given the numerous motives behind foreign aid, the more interesting
question addressed by recent scholarship centers not on the question of
*whether* donors are altruistic or self-serving, but instead on *when*.
In one prominent example, @heinrich2013 shows that when recipient policy
choices have salience for donors, aid is driven by "selfish" donor
motives. Alternatively, greater awareness of poverty in the recipient
country increases "selfless" aid giving.

Bermeo (2017, 2018) goes further, arguing that because of the increasing
interconnections between industrialized and developing countries in a
globalizing world, addressing recipient needs also serves the
self-interest of donor governments. She shows this is especially true in
proportion to the strength of the connections between an industrialized
country donor and a developing country recipient.

Both Heinrich's (2013) and Bermeo's (2017, 2018) studies point to two
variables that will help to triangulate the public or rival goods
properties of foreign aid. The first is variation in the salience of
recipient countries to donor governments. We might also label this as
donor-recipient ties. These are factors that determine the returns that
donor governments get out of their aid allocation to a particular
recipient. In Heinrich's telling, this concept relates to the salience
of recipient policy deals to donors (e.g., donor foreign policy
interest), but it also corresponds with the strength of donor-recipient
connections that, for Bermeo, make responding to recipient need import
for donors as well.

While these views of the significance of donor-recipient ties seem to be
in tension, the argument made here is that recipient need helps to
resolve the tension. One example of this is donor relationships with
former colonies. Several studies show a strong pro-colonial bias in the
aid allocation decisions of donor governments, and this finding is often
attributed to the neo-colonial aspirations of donors [see
@chibaHeinrich2019]. By targeting economic assistance to former
colonies, old colonial powers are thought to exercise political
influence absent formal colonial rule. However, colonial ties also make
donor countries especially sensitive to problems rooted in
underdevelopment in their former colonies [@bermeo2017; @bermeo2018].
These ties increase the likelihood that negative spillovers due to
poverty or conflict will have an impact on the population of the former
colonizer. It also bears noting that former colonizers may have a
comparative advantage in promoting public goods in their former
colonies. This could be due to lower marginal costs that result from
greater expertise in working with the recipient government or greater
marginal benefits because of the presence of diasporas living in the
donor country that provide side-payments. Whatever the cause (and none
of the ones proposed are mutually exclusive), ex-colonies may be
important targets of foreign aid, both for reasons rooted in donor
interest and reasons rooted in recipient need. The arguement laid out
here is that the key factor determining when one or the other objective
is front and center is variation in recipient need. Greater need may
signal greater likelihood of instability and possible negative
spillovers, which will be most salient to a donor that is a former
colonizer. However, in recipients where need is low, this may create
opportunities to pursue non-development foreign policy goals, which may
be most relevant to a donor that is a former colonizer [see @bdms2009
and @heinrich2013].

This discussion points to a simple proposition, namely, that depth of
recipient need interacts with the strength of donor-recipient
connections to shape the kinds of, and returns from, goods donors get
out of their aid allocation. As recipient need deepens, the public goods
properties of aid should be more pronounced. Conversely, as recipient
need becomes more shallow, the political opportunities that giving aid
affords donors may take on greater salience, giving way to rival goods
as donors vie for influence. Donor ties with recipients may in turn
determine the returns that donors receive from their aid allocation.

These dynamics in turn should influence donor behavior with respect to
each other. Depending on the depth of recipient need and strength of
donor ties with a recipient, donor governments will have differing
strategic incentives in how they respond to the aid giving of peers in a
particular developing country. In some recipients, incentives to compete
for top-donor status will be substantial, while in other recipients,
incentives to free-ride on the giving of others will dominate donor
decisionmaking. When we expect one or the other of these incentives to
be at work will hinge on whether public versus rival goods are the
primary motive for donor giving.

Of course, this characterization of how donor-recipient ties interact
glosses over a multitude of details. But tractability is gained in
exchange for exactness. The next section continues this discussion and
lays out two alternative hypotheses for lead donorship given the
interaction of recipient need and donor-recipient ties.

# Hypotheses

The discussion up to now points to two competing hypotheses. These
competing hypotheses follow from two different scenarios, one where
donor governments cooperate in the provision of foreign aid and the
other where they fail to do so and thus engage in a mix of competition
and buck-passing depending on the goods (public or rival) in question.

Following @steinwand2015, donor cooperation implies a delineation of
spheres of influence in a rival goods context and a mitigation of
buck-passing or free-riding with public goods. This means lead donorship
should arise with the greatest frequency when and where rival goods have
disproportionate weight relative to public goods and where one donor
obtains comparatively more marginal returns relative to others. Further,
lead donorship should occur with less frequency with public goods, even
if one donor obtains more marginal returns relative to others, as donors
agree to mitigate free-riding. This may be called the "collaboration"
hypothesis stated simply as:

\bigskip

> **H1 (Collaboration)**: *Lead donorship should emerge when and where
> foreign aid provides more rival than public goods and marginal returns
> are concentrated in a single donor.*

\bigskip

Alternatively, the absence of cooperation implies an absence of clear
spheres of influence with rival goods and the presence of buck-passing
with public goods. This means that lead donorship will have a low
likelihood of arising with rival goods, even when one donor obtains
disproportionate marginal returns. However, lead donorship will have a
greater likelihood of arising with public goods, especially when one
donor obtains greater marginal returns relative to others. We may call
this the "elusive collaboration" hypothesis. The word "elusive" is used
to modify "collaboration" to signify a *systematic* pattern of
cooperation failure. If donors fail to cooperate, but this is localized
to only some recipient countries, the elusive collaboration hypothesis
will fail to obtain. Collaboration must truly be elusive/systemic for
the following pattern to hold:

\bigskip

> **H2 (Elusive Collaboration)**: *Lead donorship should emerge when and
> where foreign aid provides more public than rival goods and marginal
> returns are concentrated in a single donor.*

\bigskip

Each hypothesis implies different patterns in lead donorship across
developing countries, which may be difficult to intuit. Figures 1 and 2
aid in the description of these patterns. Figure 1 shows the variation
in lead donorship that we would observe with donor collaboration. The
x-axis shows variation in the level of recipient need, and by symmetry
inverse variation in opportunity for using aid to realize
non-development goals. The y-axis shows variation in the degree to which
recipient ties with donors are relatively even or else concentrated in a
single donor. Under collaboration, lead donorship occurs with greatest
frequency in the upper-left quadrant of the figure where recipient need
is low and a single donor obtains greater marginal returns from private
goods relative to other donors.

```{r fig.height=4, fig.width=4, fig.cap="(Collaboration) Expected variation in lead donorship if donor governments collaborate in the provision of foreign aid. Lead donorship should be concentrated where recipient need is lowest and a single donor has disproportionate marginal returns."}
library(metR)
library(geomtextpath)
expand_grid(
  p1 = seq(0, 1, len = 10),
  p2 = seq(0, 1, len = 10)
) %>%
  mutate(
    y = (1 - p1) * p2
  ) -> dt
dy <- Derivate(y ~ p1 + p2, data = dt)
dt %>%
  mutate(
    dy.p1 = dy$y.dp1,
    dy.p2 = dy$y.dp2
  ) %>%
  ggplot() +
  aes(x = p1, y = p2, z = y) +
  geom_contour_fill() +
  geom_arrow(
    aes(dx = dy.p1, dy = dy.p2),
    color = "white"
  ) +
  geom_textabline(
    slope = -1, intercept = 1,
    label = "Increasing Likelihood of Lead Donorship",
    color = "white",
    text_only = T,
    fontface = "bold",
    vjust = 1
  ) +
  ggpal("diverging", "fill", 0.5) +
  scale_x_continuous(
    breaks = 0:1,
    labels = c("High Opportunity\nLow Need",
               "Low Opportunity\nHigh Need")
  ) +
  scale_y_continuous(
    breaks = 0:1,
    labels = c("Even Returns",
               "Concentrated Returns")
  ) +
  theme(
    axis.text.x = element_text(
      hjust = 0:1
    ),
    axis.text.y = element_text(
      angle = 90, hjust = 0:1
    ),
    legend.position = "none"
  ) +
  labs(
    x = NULL,
    y = NULL
  )
```

Figure 2 shows variation in lead donorship absent donor collaboration.
Like with Figure 1, the x-axis shows variation in recipient need and
conversely opportunity in realizing non-development or rival goods. The
y-axis shows the degree to which marginal returns are disproportionately
enjoyed by a single donor or spread more evenly across donors. Here, if
collaboration eludes donors, lead donorship will have the greatest
likelihood of occurrence where recipient need is especially high and
marginal returns are disproportionately obtained by a single donor.

```{r fig.height=4, fig.width=4, fig.cap="(Ellusive Collaboration) Expected variation in lead donorship if donor governments fail to collude in the provision of foreign aid. Lead donorship should be concentrated where recipient need is greatest and a single donor has disproportionate marginal returns."}
expand_grid(
  p1 = seq(0, 1, len = 10),
  p2 = seq(0, 1, len = 10)
) %>%
  mutate(
    y = p1 * p2
  ) -> dt
dy <- Derivate(y ~ p1 + p2, data = dt)
dt %>%
  mutate(
    dy.p1 = dy$y.dp1,
    dy.p2 = dy$y.dp2
  ) %>%
  ggplot() +
  aes(x = p1, y = p2, z = y) +
  geom_contour_fill() +
  geom_arrow(
    aes(dx = dy.p1, dy = dy.p2),
    color = "white"
  ) +
  geom_textabline(
    slope = 1, intercept = 0,
    label = "Increasing Likelihood of Lead Donorship",
    color = "white",
    text_only = T,
    fontface = "bold",
    vjust = 1
  ) +
  ggpal("diverging", "fill", 0.5) +
  scale_x_continuous(
    breaks = 0:1,
    labels = c("High Opportunity\nLow Need",
               "Low Opportunity\nHigh Need")
  ) +
  scale_y_continuous(
    breaks = 0:1,
    labels = c("Even Returns",
               "Concentrated Returns")
  ) +
  theme(
    axis.text.x = element_text(
      hjust = 0:1
    ),
    axis.text.y = element_text(
      angle = 90, hjust = 0:1
    ),
    legend.position = "none"
  ) +
  labs(
    x = NULL,
    y = NULL
  )
```

These predictions are straightforward, but testing them empirically is
not. Need is a multi-dimensional concept, as is the strength of
donor-recipient ties. After first detailing the measurement strategy for
lead donorship, the section that follows describes a novel empirical
strategy to measure need and ties. It is one that comes with some costs
to precision, but the return is a more tractable analysis that permits a
macro-level evaluation of variation in lead donorship.

# Measuring Lead Donorship

```{r}
dt <- read_csv(
  here::here("01_data", "final_data.csv")
)
stein_dt <- haven::read_dta(
  here::here("01_data", "Steinwand_DonorRecipientDyads.4.2.dta")
)
stein_dt <- stein_dt  |>
  select(donor, recipient, year, lead5of9r2)
dt <- left_join(
  dt, stein_dt, by = c("donor", "recipient", "year")
) |>
  mutate(
    lead5of9r2 = replace_na(lead5of9r2, 0)
  )
roll_ave <- function(x) {
  x <- xn <- c(NA, NA, x)
  for(i in 3:length(x)) {
    x[i] <- mean(xn[(i-2):i], na.rm=T)
  }
  x <- x[3:length(x)]
  return(x)
}
stein_ave <- function(x) {
  x <- xn <- c(NA, NA, NA, NA, x, NA, NA, NA, NA)
  for(i in 5:(length(x)-4)) {
    x[i] <- (mean(xn[(i - 4):(i + 4)], na.rm=T)>=(5/9))+0
    x2 <- 1-any(cumsum(diff(cumsum(1-xn[(i - 4):(i + 4)])))>=2)
    x[i] <- (x[i]*x2)+0
  }
  x <- x[5:(length(x)-4)]
  x
}

## Define lead donorship bilaterally:
dt %>%
  mutate(
    ties = - ties,
    need = - need
  ) %>%
  group_by(recipient, year) %>%
  mutate(
    aid_share = aid / sum(aid),
    HI = sum(aid_share^2),
    top_donor = (aid==max(aid)),
    top_share = max(aid_share),
    aid_1st = max(aid),
    aid_2nd = sort(aid, decreasing = T)[2],
    aid_diff = aid_1st - aid_2nd,
    ## aid_share = aid / sum(aid),
    ## aid_share_gt = aid_share > median(aid_share),
    ## top_donor = (aid==max(aid))+0,
    ## fst_aid = max(aid),
    ## snd_aid = sort(aid, decreasing = T)[2],
    ties_share = (ties - min(ties)) / sum(ties - min(ties))
  ) %>%
  ungroup %>%
  
  ## Steinwand's measure
  mutate(
    lead_donor_year = ifelse(
      (HI > median(HI)) &
        (aid_diff > median(aid_diff)) &
        (aid_share > median(aid_share)) &
        top_donor,
      1, 0
    )
  ) %>%
  arrange(year) %>%
  group_by(donor, recipient) %>%
  mutate(
    lead_donor = roll_ave(lead_donor_year) %>%
      ifelse(. == 0.5, 1/3, .), ## fix issue where new lead donors get score of 1/2 in second year
    lead_donor_stein = stein_ave(lead_donor_year)
  ) %>%
  ungroup -> dt
dt %>%
  mutate(
    donor_code = countrycode::countrycode(
      donor, "country.name", "iso3c"
    )
  ) %>%
  filter(
    recipient %in% c("Mexico", "Mozambique")
  ) %>%
  group_by(recipient, year) %>%
  transmute(
    donor_code = donor_code,
    aid_share = aid / sum(aid),
    aid_1st = aid_share == sort(aid_share, decreasing = T)[1],
    aid_2nd = aid_share == sort(aid_share, decreasing = T)[2]
  ) %>%
  filter(
    aid_1st | aid_2nd
  ) %>%
  ungroup %>%
  group_split(aid_2nd) -> plt_dt
```

Building on past research, @steinwand2015 proposes the following
criteria for lead donorship:

\bigskip

1.  The Herfindahl Index (HI) for a given recipient in a given year is
    greater than the median HI for the data sample;
2.  The difference between the total amount of aid given by the top
    donor relative to the next largest donor to a recipient in a year is
    greater than the sample median of this difference;
3.  The share of aid to a recipient in a given year from the top donor
    is greater than the sample median of aid shares.

\bigskip

This measure combines elements of various techniques for quantifying the
degree to which aid flows into a single recipient are from only one
donor versus many donors. The Herfindahl Index (HI) is the most commonly
applied such measure in the aid literature [see @djankovEtAl2009;
@gehringEtAl2017; @knackRahman2007; @steinwand2015]. With respect to a
recipient $r$ in year $t$ HI is given as
$$\text{HI}_{rt} = \sum_{i = 1}^N \pi_{irt}^2, \tag{1}$$ where
$\pi_{irt}$ is the share of aid given to recipient $r$ in year $t$ that
donor $i$ is responsible for. The final measure is equal to the
probability that two randomly drawn dollars from the overall aid given
to a recipient in a given year come from the same donor. The higher this
probability, the more concentrated the donor pool. While this measure is
commonly used, it lacks specificity with respect to lead donorship. This
is why the criteria for lead donroship measure also includes the two
additional benchmarks highlighted above. The goal is to take into
account not only the degree of donor concentration, but also the
absolute and relative amount of aid given to a recipient from the top
donor. To further capture the temporal dynamic of lead donorship,
@steinwand2015 adds the following longitudinal criteria:

\bigskip

4.  A donor must have top donor status for at least 5 out of 9
    consecutive years;
5.  A donor must not drop from top donor status for more than 2
    consecutive years in that 9 year time span.

\bigskip

In devising a measure of lead donorship for this study, I adopt some,
but not all, of these decision rules. In particular, I rely on the first
three while I opt for a slightly more flexible approach to deal with the
longitudinal dimension of lead donorship. Instead of the year-based
criteria, I use a simple lagged rolling average of lead donorship:

\bigskip

> *For a given year, I calculate the proportion of times a donor meets
> the criteria specified in points 1-3 in that year and the two
> previous.*

\bigskip

With this approach, if a donor met criteria 1-3 in 1999 and 2000, but
fell short in 2001, its lead donor status would be 2/3 in 2001. This
strategy adds a fuzziness to lead donorship that Steinwand's measure
lacks. Rather than being entirely present or absent, lead donorship may
wax and wane with a certain degree of granularity. For interested
readers, the Online Appendix, available on the Review of International Organizations' webpage, includes a more in depth discussion of this
measurement approach by looking at a pair of recipient country
cases---Mexico and Mozambique.

The choice to account for the longitudinal element of lead donorship has
a practical, in addition to methodological, purpose. Namely, it provides
more variation in the concept of interest. However, to facilitate
comparisons with @steinwand2015, in robustness checks I apply his more
stringent longitudinal criteria as well.

Figure 3 reports how the yearly average of lead donorship within aid
recipients compares to the yearly average of HI. For comparison, a
measure of lead donorship based on Steinwand's more restrictive temporal
criteria is included as well. Because it involves a longer bandwidth,
the measure has no valid observations for 1996 to 1999, and then from
2011 to 2014. Interestingly, while HI---which measures the probability
that two randomly drawn dollars of aid given to a recipient originate
from the same donor---slightly declined from 1998 to 2014, lead
donorship saw an increase in recipients over this same period. This
pattern runs counter to that reported by @steinwand2015. Though he
summarizes lead donorship over a slightly different period, 1970 to
2010, even for the years that overlap with the results presented here,
we identify opposite trends in lead donorship, both for the modified
measure preferred here and a measure based on Steinwand's original
criteria. Notably, because they are more restrictive, Steinwand's
measure provides a more conservative estimate of average lead donorship
over time.

```{r}
## Define lead donorship by recipient:
dt <- dt %>%
  group_by(recipient, year) %>%
  mutate(HI = sum(aid_share^2)) %>%
  ungroup
dt <- dt |>
  mutate(
    lead5of9r2 = ifelse(
      year > 2010, NA_integer_, lead5of9r2
    ),
    lead_donor_stein = ifelse(
      year %in% 2000:2010,
      replace_na(lead_donor_stein, 0),
      NA_integer_
    )
  )
dt %>%
  group_by(recipient, year) %>%
  summarize(
    lead_donor = max(lead_donor),
    HI = mean(HI),
    lead_donor_stein = max(lead_donor_stein)
  ) %>%
  group_by(year) %>%
  summarize(
    lead_donor = mean(lead_donor),
    HI = mean(HI),
    lead_donor_stein = mean(lead_donor_stein)
  ) -> year_dt
```

```{r, fig.height=3.75, fig.width=5, fig.cap="Values computed using the giving of 29 DAC countries across 24 development sectors to 127 recipient countries."}
year_dt %>%
  rename(
    "Lead Donorship" = lead_donor,
    "Lead Donorship (Steinwand)" = lead_donor_stein
  ) %>%
  ungroup %>%
  pivot_longer(
    cols = c(HI, `Lead Donorship`, `Lead Donorship (Steinwand)`)
  ) %>%
  ggplot() +
  aes(
    x = year,
    y = value,
    linetype = name
  ) +
  geom_line(size = 0.75) +
  scale_x_continuous(
    n.breaks = 10
  ) +
  labs(
    x = "Year",
    y = "Yearly Average per Recipient",
    linetype = NULL
  ) 
```

Part of the incongruence between these results may lie in the choice of
aid variable. While @steinwand2015 uses aid disbursements, I use aid
commitments. The former often closely reflects the latter but with
varying lags since commitments most closely match immediate changes in
donor policy while disbursements represent downstream realizations of
that policy over time. In a given year, Germany may commit a certain
amount of aid to Indonesia, but that aid may be disbursed in
installments over several years, for instance.

An additional difference is my choice to use the OECD's Creditor
Reporting System (CRS) to isolate aid dollars committed specifically
toward the 24 development sectors defined by the OECD. The total package
of aid donors give often includes funds for overhead, staffing,
administration, etc. I exclude such costs since they can be highly
idiosyncratic to donors and across recipients and do not capture the
types of budget support, debt relief, and sector-specific resources that
have greatest material relevance for recipient outcomes. That
@steinwand2015 observes a decline in lead donorship over time may follow
from the inclusion of overhead in aid totals---which will have increased
across donors over time as aid agencies and start-up costs for projects
grow. Including these funds, then, may have obscured the degree to which
lead donorship across development sectors has, in fact, increased over
time.

With a measure of lead donorship in hand, the next section describes the
method for measuring donor-recipient ties and need.

# Measuring Need and Donor-Recipient Ties

There are many measures that capture variation in donor interest in
particular recipients and level of development need in developing
countries. When examining only a small sample of studies, the count of
unique covariates is staggering. @bdms2009 and @berthelemyTichit2004 use
15 different measures each. @schraederEtAl1998 use over 20 and
@mckinlayLittle1979 use over 30.

Including so many variables in a regression analysis is not difficult.
The harder task is interpreting so many coefficients, each of which
reflects only a partial relationship in the data. For these reasons,
scholars like @achen2002 complain and @pearl2009 warn of the increasing
complexity of causal relationships and room for potential biases in
regression analysis that arise with each new covariate added to the
model.

The difficulties only worsen when incorporating interactions, as is the
goal of this study. Even if measures of recipient need and
donor-recipient ties could be reduced to just a handful of covariates,
because both the collaboration and elusive collaboration hypotheses
imply an interaction between these buckets of variables, the number of
n-order interactions that would be required would strain our ability to
draw sensible inferences and, worse yet, eat up valuable degrees of
freedom in exponential fashion.

The empirical challenge, then, is not the identification of variables.
Instead, the challenge lies in finding a satisfactory way of collapsing
them along the two dimensions of interest---donor-recipient ties and
recipient need.

However, before proceeding, a caveat on the value of collapsing measures
bears noting. While dimensionality reduction offers several advantages,
it is not an appropriate choice in all circumstances. While some
questions in international aid are truly about macro-level issues and
thus could benefit from synthesizing a fewer number of measures out of
many, other questions center on particular mechanisms that drive
patterns in aid giving. For example, @bermeoLeblang2015 in their study
test hypotheses linked to donor immigration policy and migrant political
mobilization as factors driving foreign aid to migrant-sending
countries. Aggregate variables would have been too blunt an instrument
for testing these mechanisms.

By contrast, composite measures are valuable in this study given the
focus on identifying general patterns in lead donorship given the
interaction of recipient need and donor-recipient ties as singular
concepts. By first reducing each to two variables, it is more tractable
to quantify how their interaction influences donor behavior.
Specifically, these measures make it possible to triangulate contexts
where motives for donor giving will be primarily rooted in genuine
interest in development promotion or in opportunistic efforts to
exercise political influence to obtain policy deals and concessions from
developing countries.

There are many approaches to dimensionality reduction, some of which are
more or less appropriate for the task at hand. Shrinkage approaches,
such as LASSO, can be tuned to eliminate variables that have little
predictive power with respect to an outcome from a model, and methods of
feature projection can be used to combine a set of covariates into a
single measure. An index could alternatively be constructed. The polity
index, for example, is a composite of several component measures such as
regime durability, regulation of chief executive recruitment, political
competition, and so on [@marshallGurr2020].

The method used here can be thought of as a blend of the feature
projection and index construction approaches. Using an objective
function, it finds the optimal set of weights with respect to both
direction and magnitude to construct a linear combination (or index)
given a set of component measures. The final product is optimized to
maximally covary with its components, appropriately weighted on the
basis of their variation. The objective function used in optimization is
the sum of the squared covariances (SSC) of each variable with a
possible linear combination of all the variables. Readers interested in
the technical details and a deeper discussion of the approach are
welcome to read the summary in the Online Appendix. The gist is that SSC provides
advantages in constructing the composite measures of donor-recipient
ties and development need relative to more familiar methods like
principal components analysis (PCA) or multivariate factor analysis
(MFA). The first is too atheoretical, optimizing on total variance
rather than balancing the need to create a linear combination that is
maximally correlated with its components. The second is too sensitive to
variable correlations with each other, which means the linear
combination may be overly influenced by the most strongly correlated
components, leaving variation due to other theoretically important
variables on the table. As is also summarized in the discussion in
the Online Appendix, SSC yields measures of the concepts of interest with greater
predictive power than versions constructed via PCA and MFA. This
superiority in terms of explanatory power is demonstrated via both
within and out-of-sample predictions and using various modeling
techniques, ranging from parametric linear regression to random forests
and neural networks.

With a method for measuring the concepts of interests identified, the
next challenge is to identify the appropriate component variables. The
concepts of donor-recipient ties and recipient need have some well
established sets of measures that are used to proxy for each. Of course,
the universe of available variables is quite large, and it would be
excessive to include all possibly relevant measures. Instead, the
selection of variables is narrowed down to four in the case of ties and
five in the case of need. These choices certainly are not the only valid
options, but they capture many of the most salient dimensions thought to
correspond to each concept. Taken together, these choices reflect common
measures used in the aid literature and capture a wide range of signals
related to interest- and needs-based drivers of foreign aid.

The need measure is composed of the following sub-measures:

\bigskip

-   yearly per capita gross domestic product (GDP);
-   yearly population size;
-   the yearly number of individuals killed due to natural disasters;
-   an indicator for whether the recipient is experiencing a civil war
    in a given year;
-   the yearly level of political and civil liberties of a recipient.

\bigskip

The first two measures were drawn from version 9.1 of the Penn World
Table (Feenstra, Inklaar, and Timmer
2015),\footnote{Available at https://doi.org/10.15141/S50T0R.} and were
log-transformed to normalize values. The third, which captures the
severity of natural disasters in a given year, is drawn from the
Institute of Health Metrics and Evaluation’s database on natural
disaster deaths reported by countries in a given year
(2021).\footnote{Available at https://ghdx.healthdata.org/gbd-results-tool.}
Values are transformed via the inverse hyperbolic sine rather than the
natural log to retain zero values. The indicator for civil war is drawn
from the UCDP/PRIO armed conflict database (Gleditsch et al. 2002;
Pettersson et al.
2021).\footnote{Available at https://ucdp.uu.se/downloads/index.html.}
It takes the value 1 for all years where there was a violent conflict
between at least two parties that involved the deaths of at least 25
combatants and which included the government as at least one of the
parties in the conflict. The final measure, recipient civil and
political liberties, is the sum of the Freedom House’s political rights
and civil liberties scores for a given recipient country in a given
year.\footnote{Available at https://freedomhouse.org/.} The rights and
liberties scores each range from 1 to 7, with higher values denoting
more violations. After summing the values, the scores were reversed so
that higher values denote greater freedom.

Each of these measures has been studied in the foreign aid literature,
and while each has an obvious correspondence with recipient need, it
bears noting that not all scholars are in agreement on whether these
factors capture recipient need or donor interests. GDP per capita, which
is often treated as a proxy for average income, has long been a variable
of interest to political scientists and economists studying donor giving
because of its obvious connection to recipient need. Early studies
looking to determine whether major donors such as the United States
allocated aid on the basis of need treated the predictive power of GDP
per capita as a telling indicator of altruistic donor motives
[@mckinlayLittle1977 @mckinlayLittle1979]. Basically, if donors give
more aid to poorer countries this should indicate that donors are
responsive to recipient need. However, others contend that economic size
may also correspond to the price that donors need to pay to extract
political concessions through aid allocation, with concessions being
cheaper from countries with poorer economies [@bdms2009]. By and large,
the aid literature consistent finds an inverse relationship between aid
and GDP per capita [@bermeo2017].

Population also is a variable that has long been of interest to aid
scholars and whose correspondence with recipient need is obvious.
However, the question of how to interpret its relationship with donor
giving has been the subject of some debate. As an indicator of recipient
need, we should expect, all else equal, a larger country to have more
need relative to an otherwise similar but smaller country [@bermeo2017].
As a result, many have used the predictive power of population as a
driver of donor giving as a test of donor motives. However, some argue
that population might also be tied to donor's strategic interests as
well. For instance, policy concessions might be cheaper to buy from
smaller countries [@bdms2009]. Alternatively, donors may see value in
gaining access to larger markets. Research on the whole shows that
donors target more aid in larger countries supporting both the idea that
donors are motivated to respond on the basis of recipient need as well
as the idea that donors have an interest in using aid to get access to
larger markets. However, some argue that when considered in the broader
context of the determinants of donor giving, population likely is
primarily an indicator of relative need [@bermeo2017].

Natural disasters are an obvious indicator of recipient need, with the
number of individuals killed by disasters a reasonable proxy for
severity. @cheng2021keeping recently use natural disasters as an
exogenous measure of recipient need that motivates a donor response. As
the authors note, immediately following a natural disaster "services
like electricity, gas, water and telecommunications may all be
disrupted" (945). In many developing countries, such disruptions can be
especially severe if the government lacks the capacity to quickly and
effectively restore power and other essential services. Therefore, more
disaster deaths should indicate deeper need for donor action.

Civil war, similarly, is a good proxy for recipient need because, as
some scholarship such as that by @collier2003breaking argues, civil war
is effectively development in reverse. Violent conflict, because of its
destructive potential, can set back economic and social development
within a country by years or possibly decades. As a result, the presence
of an ongoing civil war serves as a useful identifier of the need for a
donor response to help offset any setbacks to recipient development
caused by a violent conflict.

Finally, the presence (or absence) of political and civil liberties can
be an important contributor to recipient need. Whether such factors are
causally related to economic development is the subject of much debate
captured by a literature far to fast to cite here, it is an empirical
fact that democratic countries enjoy, on average, higher levels of
economic development. This may in part result from some factors
associated with more repressive and undemocratic regimes that may hurt
development such as poor respect for property rights, higher likelihood
of public unrest in response to government crackdowns, and so on. In
short, poor political and civil liberties can on net be yet another
indicator to donors of the level of development need in a particular
recipient country.

Taken together, these five measures contribute to an overall picture of
the relative need developing countries have for economic assistance.
They also relate in important ways. Level of income surely should matter
to donors, but the size of the country contextualizes the relative
importance of income. Conflict, too, not only contextualizes poverty,
but can in many ways be a driver of it and be driven by it. Limited
freedoms further directly play into both poverty and conflict, and vice
versa. Natural disasters, meanwhile, are exogenous, but how many people
are killed by them is both a function of total population size and a
country’s ability to adequately respond. For all these reasons,
estimating separate partial relationships in a regression analysis for
all these variables poses obvious challenges for interpretation and may
result in unreliable signs and magnitude of coefficients. But, by
combining these measures into a single composite variable, we can gain
greater insight into how the movement of need in its various
manifestations predicates the public or rival goods donors obtain from
their aid allocation.

The measure of ties, meanwhile, is constructed from the following four
variables:

\bigskip

-   bilateral distance (in kilometers) between a donor and a recipient;
-   bilateral trade (in dollars) between a donor and a recipient;
-   an indicator for whether the donor and recipient are formal allies;
-   an indicator for whether the donor and recipient share a colonial
    past.

\bigskip

The first and second measures were taken from CEPII. Distance comes from
CEPII’s gravity dataset (Mayer and Zignago 2011), and trade comes from
CEPII’s TRADEHIST dataset (Fouquin and Hugot
2016).\footnote{Both available at \url{http://www.cepii.fr/CEPII/en/bdd_modele/bdd_modele_item.asp?id=32.}}
The former is log-transformed, while the latter is transformed via the
inverse hyperbolic sine. The alliance measure is drawn from the ATOP
database and takes the value 1 if the donor and recipient share an
alliance (Leeds et al.
2002).\footnote{Available at http://www.atopdata.org/.} The colony
measure comes from the same CEPII dataset as the bilateral distance
measure, and takes the value 1 if the donor was a former colonizer of
the recipient.

Study of the role of such strategic factors highlighted above has a
well-established tradition in the literature. Extensive research has
been done to test the range of foreign policy objectives that motivate
foreign aid giving [@bearceTirone2010; @bermeo2017; @bermeo2018;
@kilbyDreher2010; @kisanganiPickering2015; @roundOdedokun2004; and
@vanDerVeen2011]. These studies and others usually subdivide donor
interests under various umbrellas. The four measures included in the
construction of ties capture four specific dimensions of donor interest:
geographic, material, strategic, and social.

Donors tend to give more aid to countries that are geographically
closer, which suggests an interest in promoting influence within their
sphere of influence. At the same time, proximity can make donors
especially vulnerable to spillovers of developing country problems
[@bermeo2017; @bermeo2018]. Distance of course is inversely proportional
to trade, a measure that captures a donor’s material or economic
interests. The correspondence between distance and trade may explain why
regression analyses that include both measures have yielded conflicting
results about the significance of
distance.\footnote{Bueno de Mesquita and Smith (2009) fail to identify a significant coefficient on distance once they control for trade, while Bermeo (2017) does not observe such a reversal of statistical significance.}
Alliances meanwhile reflect geostrategic objectives and security
commitments between donors and recipients. Research on the specific role
of alliances in predicting aid has only received limited attention from
scholars, and then only in the context of total aid
flows.\footnote{See Morgan and Palmer (2000) for an example.} However,
the existence of signed treaty agreements between donors and recipients
certainly provides material evidence of strategic interests. This
implies a role for aid in helping donors provide their allies assurances
of their security commitments. At the same time, if an ally suffers due
to instability or natural disasters, this increases the salience of
responding to recipient need. Colonial status, finally, captures
historical ties between donors and recipients. Former colonies may not
only be targeted by donors as a means to perpetuate influence, but also
because of the enduring social connections between colony and colonizer
(Alesina and Dollar 2000; Bermeo 2017).

Of course, different measures might have been used to capture need and
ties. Further, some measures associated with need---like political
freedoms or civil war---could be alternatively argued to relate to donor
interest. With respect to variable selection, these variables were
chosen on the basis of availability and judgments about their ability to
capture as much of a particular dimension of their associated concept as
possible. For example, for geostrategic interests, measures other than
alliances were considered such as arms transfers or military aid.
However, the first did not provide as much coverage as alliance
data---it contained invalid measures for a number of dyad years for
which I do have valid measures for alliances. The second, meanwhile, is
a measure that only exists in greatest detail and reliability for the
United States. While some have used US military aid to proxy for the
geostrategic interests of other donor governments (Bermeo 2017), there
are advantages to using measures that are, at minimum, unique to the
donor-recipient level.

The choice to limit measures of donor interest to bilateral factors
further relates to the decision to categorize variables like civil war
and freedom as needs-based factors rather than interest-based. While the
latter in particular has been argued to relate to donor foreign policy
goals---such as supporting the democratic regimes of developing
countries (Alesina and Dollar 2000)---it, and civil war as well, also
provides substantive information about development need. Civil wars have
been called by some as "development in reverse" because of the
detrimental effects conflict can have on economies, security, and
political stability [@stojek14]. Civil and political freedoms,
meanwhile, have been linked with long-term economic growth and societal
development [@acemogluEtAl2019]. These measures, then, capture important
signals about the overall development need of developing countries.
Further, since they are measured at the recipient, rather than
bilateral, level, it is hard to assess the unique signals they would
capture about the interests of specific donor governments. As other
scholars have argued, interest-based considerations imply that donor
governments make their aid allocation decisions on the basis of their
unique relationships with different recipient
countries.\footnote{See Berthelemy (2006).} Furthermore, at the same
time that factors like democracy or conflict signal need, their
correspondence with strategic goals actually strengthens the claim that
low need (e.g., democratic governance and peace) may signal greater
opportunity for donors to realize non-development foreign policy
interests. The U.S. has long recognized the value of stable democracy
abroad for its national
interest.\footnote{As emphasized in a recent \textit{Congressional Research Service} report: https://sgp.fas.org/crs/row/R44858.pdf.}

The measures of ties and need were constructed using data from 1995 to
2014, overlapping with the period covered by the CRS data used to
capture lead donorship. Even though the CRS data are available for more
recent years, covariates like civil war were only available up to the
year 2014. Each measure was constructed by converting each of the
relevant components to z-scores. The final linear combination produced
using SSC was then normalized to reflect z-scores as well. That means
that a value of -1 for need means that need for a particular recipient
is 1 standard deviation below average need in the sample. The
distribution of each is shown in Figure 4. Both follow relatively normal
distributions, though ties has a somewhat left-skewed distribution.

```{r fig.height=4, fig.width=8, fig.cap="Distribution of donor-recipient ties and recipient need."}
# make ties and need measures
stand <- function(x) (x - mean(x)) / sd(x)
hat_Z <- function(X, w) {
  # standardize values
  X <- apply(X, 2, function(x) stand(x))
  
  # the linear combination
  Z <- X %*% w
  
  # the the squared covariances
  covs.sqrd <- apply(X, 2, function(x) cov(x, Z)^2)
  
  # return the negative of the sum of the squared covariances
  opt <- -sum(covs.sqrd)
  return(opt)
}

# implement the objective (returns ssc)
find_Z <- function(X) {
  out <- optim(
    fn = hat_Z,
    par = rep(0, len = ncol(X)),
    X = X
  )
  Z <- X %*% out$par
  Z <- stand(Z)
  return(Z)
}

dt <- dt |>
  mutate(
    ties2 = find_Z(
      cbind(log(dist), ihs_trade, colony, ally)
    ),
    need2 = find_Z(
      cbind(log(income), asinh(pop), ihs_disaster, civilwar, fh_total)
    )
  )


dt %>%
  select(ties2, need2) %>%
  rename(
    Ties = ties2,
    Need = need2
  ) %>%
  pivot_longer(
    everything()
  ) %>%
  ggplot() +
  aes(x = value) +
  geom_histogram(
    color = "black",
    fill = "gray"
  ) +
  facet_wrap(~ name, scales = "free") +
  labs(
    x = NULL,
    y = "Count"
  ) +
  scale_y_continuous(
    labels = scales::comma
  )
```

Figures 5 and 6 summarize the correlation between the measures of ties
and need with their respective components. A correlation matrix is
shown, where blue indicates a negative correlation and red a positive
correlation. Shading is used to highlight the relative strength of the
correlation. The bivariate correlation coefficient is shown as well.
Figure 5 summarizes the correlations for the ties measure. The
correlations are not only generally strong, but they run in the
appropriate direction for the concept of donor-recipient ties. Stronger
ties correspond with more bilateral trade, less geographical distance,
sharing military alliances, and sharing a colonial past.

```{r fig.height=4, fig.width=4, fig.cap="Correlation matrix for the composit measure of ties created via SSC and its components."}
library(GGally)
# dt %>%
#   select(ties, ihs_trade, ihs_dist, ally, colony) %>%
#   cor() -> ties_cor
ties_dt <- dt %>%
    select(ties2, ihs_trade, dist, ally, colony) |>
  mutate(dist = log(dist))
names(ties_dt) <- c(
  "Ties", "Trade", "Distance",
  "Allies", "Colony"
)
ggcorr(
  ties_dt,
  label = T
) +
  ggpal("diverging", "fill") +
  theme(legend.position = "none")
```

Figure 6 shows correlations for need. Aagin, the measure captures
variation in the components in the expected direction. Greater need
corresponds with less GDP per capita (Income), greater population, more
severe disasters, lower quality of democracy, and the presence of civil
conflict.

```{r fig.height=4, fig.width=4, fig.cap="Correlation matrix for the composit measure of need created via SSC and its components."}
# dt %>%
#   select(ties, ihs_trade, ihs_dist, ally, colony) %>%
#   cor() -> ties_cor
need_dt <- dt %>%
    select(need2, income, pop, ihs_disaster, fh_total, civilwar) %>%
  mutate(income = log(income),
         pop = log(pop))
names(need_dt) <- c(
  "Need", "Income", "Population", 
  "Disaster", "Democracy", "Conflict"
)
ggcorr(
  need_dt,
  label = T
) +
  ggpal("diverging", "fill") +
  theme(legend.position = "none")
```

# Research Design

To test the collaboration and elusive collaboration hypotheses,
variation in lead donorship is modeled as a joint function of recipient
need and the concentration of donor-recipient ties. The analysis is done
at the recipient-year level. The following empirical model is estimated:
$$\text{logit}(\text{Lead Donorship}_{rt}) = s(\tau_{rt}, \nu_{rt}) + \alpha_t + \rho_r + \epsilon_{rt}. \tag{2}$$

The outcome is the strength of lead donorship in a developing country
for the top donor. The logit transformation is there to denote that the
model is estimated using a quasi-binomial logit link using maximum
likelihood (e.g., a fractional logit). This helps to ensure that fitted
values for lead donorship remain constrained to the unit interval.

The function $s(\cdot)$ is a multivariate nonlinear smoother that
permits estimating the joint, interactive relationship between lead
donorship and the measures of interest. In particular, a thin plate
regression spline is used. This particular smoother has a number of
documented advantages over several alternatives including computational
efficiency, avoiding problems associated with knot placement, and a
sensible way to model interactions [@wood2003thin]. This more flexible
non-linear approach is adopted because the goal is to test not a simple
directional linear hypothesis, but instead a pair of competing conjoint
conditions under which lead donorship will emerge. A multivariate
smoother is not only sufficient for this task, it has the advantage of
not being sensitive to the parametric assumptions that would attend a
more conventional linear regression with a second order interaction
term. For more information about the model used here and about
generalized additive models in general, interested readers should refer
to @wood2020inference.

The multivariate smoother has two variable inputs: $\tau_{rt}$ and
$\nu_{rt}$. The first is a measure of donor tie concentration. After,
first, using the SSC method outlined in the earlier discussion to create
a measure of donor-recipient ties, an HI index of the ties variable was
created at the recipient-year level. After log-transformation to
normalize the data, the resulting measure, denoted $\tau_{rt}$, captures
the degree to which ties between a recipient a donor country are
concentrated in a single donor. Higher values equal greater
concentration. The measure $\nu_{rt}$ denotes the level of recipient
need as constructed via SSC. Higher levels equal greater development
need. To guard against overfitting while ensuring sufficient flexibility
to handle nonlinearities in the data, $k = 15$ degrees of freedom is set
as the upper bound for the smoother.

The model contains two additional terms. $\alpha_t$ captures year
specific intercepts that adjust for unobserved year-to-year shocks. This
ensures that variation in lead donorship reflects within-year variation
rather than across years. The second term, $\rho_r$, denotes recipient
random effects that adjust for unobserved recipient level heterogeneity
and dependence over time. Finally, the model is rounded off with a
zero-mean error term.

The data used to fit the model consists of 2,398 unique recipient-year
observations from 1997 to 2014. These are comprised of a total of 133
unique developing country aid recipients.

The analysis in the next section proceeds as follows. Before summarizing
the main results, a dyadic analysis of donor aid commitments is done to
provide some face validity for the measures of donor-recipient ties and
recipient need. If the measures prove to be poor determinants of dyadic
aid flows, this would raise doubts about their ability to help
triangulate the publicness or rivalness and relative marginal returns of
goods donors get from their aid allocation. A summary of the main
analysis follows.

# Analysis

## Face Validity

Table 1 summarizes regression coefficients from five linear models. Each
was estimated with OLS with standard errors clustered by dyad. All
include year and donor fixed effects. They were estimated using a total
of 68,856 dyad-year observations.

```{r results='asis'}
ols_fit1 <- lm_robust(
  ihs_aid ~ 
    ## dyadic factors
    ihs_trade + ihs_dist + colony + ally +
    ## recipient factors
    log(income) + log(pop) + ihs_disaster + civilwar + fh_total,
  data = dt,
  fixed_effects = ~ year + donor,
  se_type = "stata",
  clusters = paste0(recipient, donor)
)
ols_fit2 <- lm_robust(
  ihs_aid ~ 
    ## dyadic factors
    ties2 +
    ## recipient factors
    log(income) + log(pop) + ihs_disaster + civilwar + fh_total,
  data = dt,
  fixed_effects = ~ year + donor,
  se_type = "stata",
  clusters = paste0(recipient, donor)
)
ols_fit3 <- lm_robust(
  ihs_aid ~ 
    ## dyadic factors
    ihs_trade + ihs_dist + colony + ally +
    ## recipient factors
    need2,
  data = dt,
  fixed_effects = ~ year + donor,
  se_type = "stata",
  clusters = paste0(recipient, donor)
)
ols_fit4 <- lm_robust(
  ihs_aid ~ 
    ## dyadic factors
    ties2 +
    ## recipient factors
    need2,
  data = dt,
  fixed_effects = ~ year + donor,
  se_type = "stata",
  clusters = paste0(recipient, donor)
)
ols_fit5 <- lm_robust(
  ihs_aid ~ 
    ## dyadic factors
    ties2 +
    ## recipient factors
    need2 +
    ## interaction
    ties2:need2,
  data = dt,
  fixed_effects = ~ year + donor,
  se_type = "stata",
  clusters = paste0(recipient, donor)
)
texreg(
  list(ols_fit1, ols_fit2, ols_fit3, ols_fit4, ols_fit5),
  include.ci = F,
  caption = "OLS Estimates for Dyadic Aid Commitments",
  caption.above = T,
  custom.header = list("Bilateral ODA Commitments" = 1:5),
  custom.coef.names = c(
    "Trade (asinh)",
    "Distance (asinh)",
    "Colony",
    "Alliance",
    "Income (log)",
    "Population (log)",
    "Disaster (asinh)",
    "Civil War",
    "Democracy",
    "Ties",
    "Need",
    "Ties $\\times$ Need"
  ),
  custom.gof.rows = list(
    "Year FE" = rep("Yes", len = 5),
    "Donor FE" = rep("Yes", len = 5),
    "N" = rep(scales::comma(nrow(dt)), len = 5)
  ),
  include.nobs = F,
  include.nclusts = F,
)
```

Model 1 was estimated using each of the separate component measures used
in the construction of ties and need. Model 2 replaces the separate
measures for ties with the composite ties measure. Model 3 does the
same, but for the need measure. Model 4 includes only the composite
measures while excluding the components. Model 5 adds an interaction
between ties and need.

As would be expected with any method of dimensionality reduction, the
R-squared value slightly worsens when the component measures are
replaced by their composites. However, the reduction in variance
explained is not substantial. Indeed, given the reduction from nine
covariates to two, the composite measures explain an impressive amount
of variation in dyadic aid commitments.

The direction of the estimates is further consistent with what we would
expect. Both stronger bilateral ties and greater recipient need
correspond with greater bilateral aid commitments from donors. The
results for Model 5 are especially interesting. There is a statistically
significant and substantial positive interaction between donor-recipient
ties and recipient need. This pattern is consistent with recent
arguments that donor governments have additional incentives to respond
to recipient need when they share stronger bilateral connections
[@bermeo2017; @bermeo2018]. Taken together, these results are
encouraging for the main analysis because they provide some face
validity for the composite measures. Not only are they strong predictors
of dyadic aid flows, but they also interact in accordance with recent
arguments about donor behavior.

## Predicting Lead Donorship

The analysis now turns to the main results. Recall that the unit of
observation is the recipient-year and the outcome of interest the
strength of lead donorship for the top donor in a recipient. Table 2
reports estimates for two fractional logistic models with
semi-parametric smoothers for the measures of ties concentration and
recipient need. In one, the measure of ties and of need were given as
inputs to separate smoothing functions. In the other, they were inputs
to a joint smoothing function. Models were estimated using year fixed
effects and recipient random effects. Values reported are not regression
coefficients but effective degrees of freedom. The statistical
significance of each corresponds to the overall quality of the fit for
the smoother function. The magnitude of the EDF corresponds to the
degree of nonlinearity of the smoothers. An EDF of 1 would imply a
linear relationship while values of increasing magnitude indicate
nonlinear relationships. The values reported in Table 2 imply a high
degree of nonlinearity.

```{r}
## Aggregate the data
hhi <- function(x) {
  tot <- sum(x, na.rm=T)
  pct <- x / tot
  ssp <- sum(pct^2, na.rm=T)
  ssp # return
}
dt %>%
  group_by(
    year, recipient
  ) %>%
  mutate(
    ties_share = (ties2 - min(ties2)) / sum(ties2 - min(ties2)),
  ) %>%
  reframe(
    lead_donor = max(lead_donor, na.rm=T),
    lead_donor_stein = max(lead_donor_stein, na.rm=T),
    aid_hhi = log(sum(aid_share^2, na.rm=T)),
    ties_ssp = log(sum(ties_share^2, na.rm=T)),
    need = mean(need2, na.rm=T)
  ) %>%
  mutate(
    num_dyad = as.numeric(as.factor(recipient))
  )-> agg_dt

## Estimate the GAM model
library(mgcv)
gam_fit1 <- gam(
  lead_donor ~ s(ties_ssp) + s(need) + 
    as.factor(year) + 
    s(num_dyad, bs = "re"),
  data = agg_dt,
  method = "REML",
  family = quasibinomial
)
gam_fit2 <- gam(
  lead_donor ~ s(ties_ssp, need, k = 15) + 
    as.factor(year) + 
    s(num_dyad, bs = "re"),
  data = agg_dt,
  method = "REML",
  family = quasibinomial
)
```

```{r results='asis'}
library(texreg)
texreg(
  list(Separate = gam_fit1, Joint = gam_fit2),
  custom.coef.map = list(
    "EDF: s(ties_ssp)" = "Ties Concentration",
    "EDF: s(need)" = "Recipient Need",
    "EDF: s(ties_ssp,need)" = "Joint Ties and Need"
  ),
  custom.header = list("Lead Donorship" = 1:2),
  caption = "Logistic GAM Estimates",
  caption.above = T,
  include.aic = F,
  include.bic = F,
  include.log = F,
  include.nsmooth = F,
  include.nobs = F,
  include.nclusts = F,
  digits = 3,
  custom.gof.rows = list(
    "Year FE" = rep("Yes", len = 2),
    "Recipient RE" = rep("Yes", len = 2),
    "N" = c(scales::comma(nrow(agg_dt)), 
            scales::comma(nrow(agg_dt))),
    "Recipients" = c(scales::comma(length(unique(agg_dt$recipient))),
                scales::comma(length(unique(agg_dt$recipient))))
  )
) %>% cat()
```

A comparison of the separate and joint smoothers for ties concentration
and need provides clear reason to favor a joint smoother. In model 1,
only the measure of need has a significant relationship with lead
donorship, but in model 2 the joint smoother on need and ties is
statistically significant. While Model 1 suggests that ties
concentration does no explain variation in lead donorship, Model 2 makes
clear that once ties are considered in combination with recipient need,
this factor does help explain variation in lead donorship. As shown in
the model diagnostics, with the joint smoother the model explains
greater variation in lead donorship than when need and ties are treated
seperately. This is consistent with both the collaboration (H1) and
elusive collaboration (H2) hypotheses which hold recipient need
interacts with concentration of marginal returns from aid to determine
lead donorship. However, this conclusion is as far as an examination of
Table 2 can take us. Because of the nonlinear and interactive
relationship implied by the model, visualization of the results is
necessary to see how ties and need jointly predict lead donorship.

Figure 7 summarizes these predictions. Along the x-axis is variation in
recipient need. The y-axis shows variation in the concentration of donor
ties. Across the range of possible values of each measure, the
likelihood of lead donorship is plotted. Blue denotes that the
probability of lead donorship is less than 50% while red indicates that
the probability is greater than 50%. Arrows help to highlight the
direction of an increase in the likelihood of lead donorship.

```{r fig.height=4.5, fig.width=4, fig.cap="Predicted (%) likelihood of lead donorship given recipient need and concentration of donor-recipient ties. Fitted logistic GAM estimates using recipient-year data from 1997-2014."}
ties_range <- agg_dt$ties_ssp |> range()
need_range <- agg_dt$need |> range()
pred_data <- expand_grid(
  ties_ssp = seq(ties_range[1], ties_range[2], len = 20),
  need = seq(need_range[1], need_range[2], len = 20),
  num_dyad = 1,
  year = 2001
)
pred_data %>%
  mutate(
    pred = predict(gam_fit2, newdata = ., type = "response")
  ) -> pred_data
dy <- Derivate(pred ~ need + ties_ssp, data = pred_data)
pred_data %>%
  mutate(
    dy.p1 = dy$pred.dneed,
    dy.p2 = dy$pred.dties_ssp
  ) -> pred_data
ggplot(pred_data) +
  aes(
    x = need,
    y = ties_ssp,
    z = pred
  ) +
  geom_contour_fill() +
  geom_arrow(
    aes(dx = dy.p1, dy = dy.p2),
    color = "white",
    size = 0.1,
    skip = 1
  ) +
  ggpal(
    type = "diverging",
    aes = "fill",
    breaks = c(0.25, 0.5, 0.75),
    labels = scales::percent,
    midpoint = 0.5
  ) +
  labs(
    x = expression(
      "Recipient Need (" * nu[r][t] * ")"
    ),
    y = expression(
      "Concentration of Ties (" * tau[r][t] * ")" 
    ),
    fill = "Likelihood of\nLead Donorship",
    caption = expression(""%->%" Direction of Greater Lead Donorship")
  ) +
  theme(
    legend.position = "top"
  )
```

Both the collaboration and elusive collaboration hypotheses imply
different corners of the figure in which we should see lead donorship
arise with the greatest likelihood (see Figure 1 and Figure 2 earlier in
the paper). Collaboration implies that lead donorship should emerge
disproportionately in the upper left quadrant, while an absence of
collaboration implies that lead donorship should emerge more often in
the upper right.

The results in Figure 9 leave little to the imagination. Lead donorship
clearly occurs with the greatest frequency at the intersection of both
high recipient need and strongly concentrated donor-recipient ties. This
is consistent with the elusive collaboration hypothesis. That is, the
pattern in lead donorship is what we would expect if donors fail to
cooperate in the delineation of spheres of influence with rival goods
and fail to mitigate free-riding with public goods. In short, the
results are consistent with systematic cooperation failure.

## Do the Results Hold with a More Restrictive Measure of Lead Donorship?

```{r}
## fit models with Steinwand based measure
gam_fit3 <- gam(
  lead_donor_stein ~ s(ties_ssp) + s(need) + 
    as.factor(year) + 
    s(num_dyad, bs = "re"),
  data = agg_dt |> 
    filter(between(lead_donor_stein, 0, 1)),
  method = "REML",
  family = quasibinomial
)
gam_fit4 <- gam(
  lead_donor_stein ~ s(ties_ssp, need, k = 15) + 
    as.factor(year) + 
    s(num_dyad, bs = "re"),
  data = agg_dt |> filter(between(lead_donor_stein, 0, 1)),
  method = "REML",
  family = quasibinomial
)

## fit using lagged version of ties
agg_dt |>
  group_by(recipient) |>
  mutate(
    lag_ties_ssp = lag(ties_ssp, order_by = year, n = 2L)
  ) |>
  ungroup() -> agg_dt
gam(
  lead_donor ~ s(lag_ties_ssp, need, k = 15) + 
    as.factor(year) + 
    s(num_dyad, bs = "re"),
  data = agg_dt,
  method = "REML",
  family = quasibinomial
) -> gam_fit5
```

It is important to check whether using lead donorship criteria
established by @steinwand2015 yields different results to those above.
As noted, Steinwand's measure of lead donorship differs from that used
in the main analysis in two ways. First, it sets a higher bar along the
temporal dimension. Second, it is binary. That means it is much more
conservative in assigning lead donor status and that it provides less
variation that might be exploited to test the hypotheses. How do the
results hold up with this more restrictive measure?

```{r results='asis'}
nagg_dt <- agg_dt |>
  filter(between(lead_donor_stein, 0, 1))
texreg(
  list(Separate = gam_fit3, Joint = gam_fit4),
  custom.coef.map = list(
    "EDF: s(ties_ssp)" = "Ties Concentration",
    "EDF: s(need)" = "Recipient Need",
    "EDF: s(ties_ssp,need)" = "Joint Ties and Need"
  ),
  custom.header = list("Lead Donorship (Steinwand)" = 1:2),
  caption = "Logistic GAM Estimates",
  caption.above = T,
  include.aic = F,
  include.bic = F,
  include.log = F,
  include.nsmooth = F,
  include.nobs = F,
  include.nclusts = F,
  digits = 3,
  custom.gof.rows = list(
    "Year FE" = rep("Yes", len = 2),
    "Recipient RE" = rep("Yes", len = 2),
    "N" = c(scales::comma(nrow(nagg_dt)), 
            scales::comma(nrow(nagg_dt))),
    "Recipients" = c(scales::comma(length(unique(nagg_dt$recipient))),
                scales::comma(length(unique(nagg_dt$recipient))))
  )
) %>% cat()
```

Table 3 reports the EDF for the nonlinear smoothers for logistic models
similar to the ones specified before. The difference is that the
semi-continuous measure of lead donorship is replaced with the binary
measure of lead donorship based on Steinwand's longitudinal criteria. As
before, the model includes year fixed effects and recipient random
effects. The first column shows results for separate smoothers applied
to donor tie concentration and recipient need. The second column shows
results for a joint multivariate smoother applied to both measures.
Because the new measure of lead donorship lacks valid observations for
the first four and last four years in the time-series, the sample of
recipient-year observations is smaller than in the previous analysis.

Despite these differences, an inspection of Table 3 provides results
consistent with the main analysis. While tie concentration is not a
significant predictor, but recipient need is, in the first model, the
model with a joint smoother provides a better fit than the model with
separate smoothers for need and ties suggesting that ties concentration
does explain variation in lead donorship once it's relationship with
lead donorship is allowed to be conditioned by recipient need.

```{r fig.height=4.5, fig.width=4, fig.cap="Predicted (%) likelihood of lead donorship based on Steinwand's (2015) more restrictive longitudinal criteria, given recipient need and concentration of donor-recipient ties. Fitted logistic GAM estimates using recipient-year data from 1997-2014."}
ties_range <- (agg_dt |> filter(
  between(lead_donor_stein, 0, 1)
))$ties_ssp |> range()
need_range <- (agg_dt |> filter(
  between(lead_donor_stein, 0, 1)
))$need |> range()
pred_data <- expand_grid(
  ties_ssp = seq(ties_range[1], ties_range[2], len = 20),
  need = seq(need_range[1], need_range[2], len = 20),
  num_dyad = 1,
  year = 2001
)
pred_data %>%
  mutate(
    pred = predict(gam_fit4, newdata = ., type = "response")
  ) -> pred_data
dy <- Derivate(pred ~ need + ties_ssp, data = pred_data)
pred_data %>%
  mutate(
    dy.p1 = dy$pred.dneed,
    dy.p2 = dy$pred.dties_ssp
  ) -> pred_data
ggplot(pred_data) +
  aes(
    x = need,
    y = ties_ssp,
    z = pred
  ) +
  geom_contour_fill() +
  geom_arrow(
    aes(dx = dy.p1, dy = dy.p2),
    color = "white",
    size = 0.1,
    skip = 1
  ) +
  ggpal(
    type = "diverging",
    aes = "fill",
    breaks = c(0.25, 0.5, 0.75),
    labels = scales::percent,
    midpoint = 0.5
  ) +
  labs(
    x = expression(
      "Recipient Need (" * nu[r][t] * ")"
    ),
    y = expression(
      "Concentration of Ties (" * tau[r][t] * ")" 
    ),
    fill = "Likelihood of\nLead Donorship",
    caption = expression(""%->%" Direction of Greater Lead Donorship")
  ) +
  theme(
    legend.position = "top"
  )
```

Turning to Figure 8, the predicted likelihood of lead donorship based on
the model shown in the second column of Table 3 follows a pattern
consistent with that shown in the main analysis. Notably, the results
are less crisp, which is to be expected. Also, because this measure of
lead donorship is more conservative, predicted levels of lead donorship
do not rise to the levels observed in the previous analysis. Even so,
the highest probabilities of lead donorship are still observed in
recipients that are both high need and that have bilateral ties
concentrated in one or a few donors. This pattern is again consistent
with the elusive collaboration hypothesis.

## Are the Results Driven by the Measurement Method?

A valid question to consider is whether the method (SSC) of constructing
the measure of recipient need and donor-recipient ties concentration is
responsible for driving the results. If alternative approaches (PCA or
MFA) yield different conclusions, this would be problematic. To test
whether this is the case, as a robustness check the main analysis was
repeated using PCA and MFA constructed versions of the measures of
interest. Figure 9 reports the results using PCA, and Figure 10 reports
the results using MFA.

```{r fig.height=4.5, fig.width=4, fig.cap="Predicted likelihood of lead donorship given recipient need and concentration of donor-recipient ties using PCA constructed measures. Fitted logistic GAM estimates using recipient-year data from 1997-2014.", fig.align="center"}
knitr::include_graphics(
  here::here("pca_results.png")
)
```

```{r fig.height=4.5, fig.width=4, fig.cap="Predicted likelihood of lead donorship given recipient need and concentration of donor-recipient ties using MFA constructed measures. Fitted logistic GAM estimates using recipient-year data from 1997-2014.", fig.align="center"}
knitr::include_graphics(
  here::here("mfa_results.png")
)
```

With the PCA constructed measures the same substantive conclusions are
drawn from the data as in the main analysis---we observe lead donorship
emerge at the influx of high need and high concentration of
donor-recipient ties. However, the analysis with MFA diverges from both
the SSC and PCA derived measures. Rather than seeing lead donorship
emerge at the upper right quadrant (as we do with SSC and PCA) or the
upper left, it arises with greatest frequency across the entirety of the right in only the most extreme cases of need. Correlations between the MFA measure and the
alternatives as well as the component measures show that the MFA
constructed measure of need may not do an adequate job of capturing
variation in the targeted construct. While it is strongly correlated with
the PCA and SSC versions of need, its correlations with the
components used to construct are comparatively weaker, save for the case of recipient population. In
short, the MFA based measure of need is not strongly correlated with the
component measures used to capture need, save for recipient population. Meanwhile, the PCA and SSC
versions of need are nearly identical and more strongly correlated with their component measures. This does not prove, but does
suggest, that MFA does a worse job of capturing the concept of recipient
need from the component measures. PCA, on the other hand, seems to do a
better job.

## Is It Really Cooperation Failure?

Based on the theoretical argument made by @steinwand2015, the results in
this study are consistent with systemic cooperation failure in the
allocation of foreign aid to developing countries. This cooperation
failure would seem to be characterized by a mix of donor competition for
top spot in some aid recipients and buck-passing in other recipients,
leading to a pattern of lead donorship emerging primarily in higher-need
developing countries that have bilateral ties concentrated in one or a
few donors. However, other explanations are worth exploring. Two in
particular bear special mention.

On the one hand, it could be that lead donorship arises in high-need
recipients with ties disproportionately concentrated in a single donor
due to donor coordination to limit redundancies in aid projects and to
ease burdens on recipient bureaucracy in high need developing countries.
On the other hand, it could be that the role of donor interactions in
driving patterns in lead donorship is overstated---that instead lead
donorship is epiphenomenal to the interests of donor countries. An
admitted limitation with this study is that it is difficult to tease out
these different explanations empirically. These arguments can all be
consistent with the patterns we observe in the data. However, there are
reasons to view these alternative explanations with some skepticism. Let
us consider them in order.

### Could It Be Coordination?

The first idea---that lead donorship is a function of donor coordination
rather than cooperation failure---would certainly be consistent with
some research on aid fragmentation and aid effectiveness that suggests
recipient outcomes are better when they only have to work with one or a
few donors [@kanbur2006; @knackRahman2007; @anderson2012;
@hernandez2017; @zeitz2021]. By coordinating on lead donorship, the idea
would be that the donor community is able to optimize development
outcomes in developing countries. If this is really what is going on,
then it is encouraging that we observe lead donorship emerge among many
high-need developing countries that would benefit the most from aid
effectiveness; albeit, donor coordination in this view would appear to
be limited to recipients where heterogeneous donor-recipient ties serve
as an organizing principal for which donor will assume lead donor
status.

However, for coordination to be the best explanation for the pattern of
lead donorship observed in the data, a couple of assumptions must be
met. To start, it must be that coordination in high-need recipients
really is optimal. Not all scholarship is in agreement on this issue.
Some recent evidence suggests that concerns about aid fragmentation for
aid effectiveness are overstated and that in some cases fragmentation
can actually support rather than detract from achieving development
goals [@gehringEtAl2017]. This is not only the case with respect to
development, it also seems to be true in the case of political
destabilization and conflict in aid recipients. As @gutting2017donor
find, fragmentation can actually limit the severity of so-called "aid
shocks" (sudden and severe cuts in foreign aid from a donor) and thereby
mitigate the risk of violent civil conflict.

An emerging idea in the literature is that aid fragmentation's
consequences are conditional, depeding on the type of aid given and the
recipient context [@gehringEtAl2017; @carlitz2023harmony]. The idea that
fragmentation's consequences for aid effectiveness are conditional is
actually central to the argument @steinwand2015 makes as well. In his
view, fragmentation is inefficient in the context of donor competition,
while it is beneficial in the context of public goods. Indeed, studies
that look at mechanisms for aid fragmentation often find that
competition is often to blame, as @barthelEtAl2014 and @fuchsEtAl2015
find in the case of donor interests in gaining access to export markets
and shoring up political support. In short, if coordination is to be in
the best interest of the actors involved, it would be where rival goods
are at stake rather than public goods. Thus, if lead donorship is a
function of coordination, it would make sense that it would appear where
public goods are less salient, that is, in low-need recipients where
donor-recipient ties can serve as a basis for coordinating on which
donor will assume lead donor status. This prediction is consistent with
the competing "collaboration" hypothesis proposed in this study, which
the main analysis fails to support.

A second assumption that must hold is that the level of coordination
required to allow lead donorship to arise where and when it does is
possible or, for that matter, recognized as such by actors in the
international development community. By the numbers, some non-zero level
of lead donorship exists in approximately 45% of developing
country-years in the data sample used in the main analysis. This
suggests a level of donor coordination that would be surprising to
anyone familiar with evaluations on donor progress on various
development targets set at high-level fora. As the OECD notes regarding
the history of these fora in the twenty-first century, the first of
which was in Rome in 2003, "success has not always been evident: lack of
co-ordination, overly ambitious targets, unrealistic time and budget
constraints and political self-interest have too often prevented aid
from being as effective as desired."[^1] Some observers have gone so far
as to call efforts to promote and maintain donor collaboration in
meeting effective development goals as beyond "insufficient"—these
efforts might as well be "dead."[^2] If evidence of lead donorship at
the intersection of high need and high donor-recipient ties is evidence
of effective donor coordination in many aid recipients across time, this
would be news to the community of international development actors.

[^1]: "The High Level Fora on Aid Effectiveness: A history." Accessed
    May 7, 2024.
    <https://www.oecd.org/dac/effectiveness/thehighlevelforaonaideffectivenessahistory.htm>.

[^2]: "After Geneva, the Effectiveness Agenda Still Lacks Direction."
    Accessed May 7, 2024:
    <https://www.cgdev.org/blog/after-2022-geneva-effectiveness-summit-effectiveness-agenda-still-lacks-direction>

Supposing the above assumptions can be met, there is the additional
issue that where public goods are concerned, the aid literature by and
large treats free-riding as the most likely problem to arise among
donors [@bermeo2018; @mascarenhasSandler2006; @steinwand2011;
@steinwand2015]. Just because this is a common framework for
understanding aid allocation in the context of public goods does not
imply that this is the best framework. However, free-riding does offer a
simple explanation for why lead donorship has the greatest likelihood of
arising when and where it does. Also, if lead donorship is the product
of uncooperative behavior, this is easier to square with the fact that
there is limited evidence that coordination improves aid effectiveness
with respect to many development outcomes and the fact that actors in
the international development community express chronic disappointment
about progress on donor collaboration.

### Is Lead Donorship Just a Function of Donor Interest?

A second alternative explanation for when and where lead donorship
arises is that lead donorship is epiphenomenal to the interests of donor
countries. To be sure, it is reasonable to assume that donor interest in
giving aid has a higher baseline in higher need developing countries,
and that by extension stronger ties magnify this interest. This is, in
fact, what the dyadic aid analysis in the "Face Validity" section seems
to demonstrate when donor-recipient ties are interacted with recipient
need.

This fact on its own could certainly drive some of the results in the
main analysis, even if the basic theoretical argument still holds true.
Indeed, an examination of the data shows that average and median aid
received in high-need recipients with concentrated ties is higher than
in other subsets of the sample. However, for this to be the only factor
driving lead donorship requires some additional assumptions to hold. The
first assumption is that donors by and large give aid without regard for
the giving of other donors. The second assumption is that donors derive
no benefit from the production of public goods in recipients they share
weak ties with. The third assumption is that in low-need recipients,
variation in donor-recipient ties does little to differentiate which
donors are likely to be responsible for giving the greatest aid.

The first of these assumptions seems especially tenuous. Whether the
source of donor interaction is presumed to be public goods or rival
goods, many studies have found evidence that donors react to the giving
of their peers, whether that be in the aggregate or in specific aid
recipients, and whether the reaction be positive or negative
[@barthelEtAl2014; @daviesKlasen2019; @dudley1979; @fuchsEtAl2015;
@mascarenhasSandler2006; @steinwand2011; @steinwand2015]. These patterns
could be an artifact of omitted variable bias, yet the finding that
donors react to the giving of peers remains a persistent finding across
studies.

With respect to the second assumption, it is theoretically possible for
a donor to get zero benefit from the production of public goods in a
recipient that is geographically, politically, and economically remote
from it. However, in an increasingly interconnected world, problems
rooted in underdevelopment in remote places can all-too-easily pose
negative externalities for a distant donor. Indeed, in summarizing the
proclamations of many industrialized country governments, @bermeo2018
notes a clear awareness on the part of country leaders of the threat
that poverty in even the most faraway places can pose to a donor's
national interest and the wellbeing of its domestic population. The
forces of globalization are not constant, to be sure. The whole point of
needing to measure the strength of donor-recipient ties is that donor
susceptibility to negative externalities rooted in underdevelopment is
variable. However, to say that this susceptibility is effectively zero
for a non-trivial number of donors with respect to a non-trivial number
of aid recipients is a strong assumption—one that is made increasingly
hard to maintain with the seemingly unstoppable inertia of
globalization.

The idea with the third assumption is that donors receive negligible
returns from giving aid in low-need recipients, which is why lead
donorship is less common among this subset of recipients even when
donor-recipient ties are concentrated. This is proposed as an
alternative to the claim made in the present study that donor
competition in this subset of recipients is the reason why lead
donorship does not emerge. Again, on this front, among the subset of
studies that focus on the role of donor competition, the idea that
competition does not account for the absence of a top donor in at least
a few aid recipients seems tenuous. Whether the source of competition be
generic or specific to particular donor interests like trade, a number
of studies point to donor competition playing a role in their aid
allocation decisions.

The idea that lead donorship is entirely a product of the
interdependence of donor aid allocation decisions is, of course, far
fetched. With out a doubt, donor interests likely play an independent
role in driving lead donorship as well. It would be surprising if this
were not the case. However, the converse argument that donor interests
are all that matter for determining lead donorship is difficult to
reconcile with past research and globalization.

## What about Endogeneity?

Another issue with the above analysis worth considering is whether lead
donorship is endogenous with donor-recipient ties. For example, could it
be the case that donor governments develop stronger ties with some
recipients over others because they are also lead donors? This is an
interesting question to consider because answering it in the affirmative
implies that lead donorship and donor-recipient ties are mutually
reinforcing. Such a view would not be unreasonable. In a recent survey
of the aid literature, @dreher2024aid conclude that the evidence
supports a positive effect of foreign aid on promoting various bilateral
donor interests, which certainly suggests endogeneity may be an issue
for the measure of donor-recipient ties used in the main analysis. Of
course, the question that must be answered is how much is endogeneity an
issue such that it threatens the inferences made in this study?

```{r fig.height=4.5, fig.width=4, fig.cap="Predicted (%) likelihood of lead donorship given recipient need and concentration of donor-recipient ties using a two-year lag of ties concentration. Fitted logistic GAM estimates using recipient-year data from 1997-2014."}
ties_range <- agg_dt$lag_ties_ssp |> range(na.rm = T)
need_range <- agg_dt$need |> range()
pred_data <- expand_grid(
  lag_ties_ssp = seq(ties_range[1], ties_range[2], len = 20),
  need = seq(need_range[1], need_range[2], len = 20),
  num_dyad = 1,
  year = 2001
)
pred_data %>%
  mutate(
    pred = predict(gam_fit5, newdata = ., type = "response")
  ) -> pred_data
dy <- Derivate(pred ~ need + lag_ties_ssp, data = pred_data)
pred_data %>%
  mutate(
    dy.p1 = dy$pred.dneed,
    dy.p2 = dy$pred.dlag_ties_ssp
  ) -> pred_data
ggplot(pred_data) +
  aes(
    x = need,
    y = lag_ties_ssp,
    z = pred
  ) +
  geom_contour_fill() +
  geom_arrow(
    aes(dx = dy.p1, dy = dy.p2),
    color = "white",
    size = 0.1,
    skip = 1
  ) +
  ggpal(
    type = "diverging",
    aes = "fill",
    breaks = c(0.25, 0.5, 0.75),
    labels = scales::percent,
    midpoint = 0.5
  ) +
  labs(
    x = expression(
      "Recipient Need (" * nu[r][t] * ")"
    ),
    y = expression(
      "Concentration of Ties (" * tau[r][t] * ")" 
    ),
    fill = "Likelihood of\nLead Donorship",
    caption = expression(""%->%" Direction of Greater Lead Donorship")
  ) +
  theme(
    legend.position = "top"
  )
```

Fortunately, at least two of the component variables used to construct
the measure of donor-recipient ties are plausibly exogenous: bilateral
distance and colonial past. While a donor's generosity cannot warp space
to narrow its geographic distance to a recipient, past donor
colonization of recipients is such a distant historical fact that
contemporary donor giving is not determinative of the former-colony
status of recipients today.

However, the remaining two component measures of donor-recipient ties
(trade and alliances) may suffer from an endogeneity problem. Trade in
particular is likely the most problematic. A sizable body of research
has emerged around the question of whether donors reward trade partners
with more aid and whether aid promotes greater levels of future trade
between recipients and donors. Summarizing the results from several
studies, @dreher2024aid conclude that donors both target aid to
recipients on the basis of trade and also that aid has a positive impact
on trade between recipients and donors. Though alliances also may be
endogenous to donor giving, research on the link between aid and
alliances is less well-developed. More than three decades ago,
@walt1990origins argued that aid's ability to strengthen formal military
alliances or to lead to the creation of new ones is likely limited;
although he based this claim only on a limited set of case studies and
descriptive analyses. Other empirical research at the intersection of
aid and alliances considers whether aid giving is determined by
alliances but does not test the reverse relationship [@palmer2002give;
@zhang2023does]. This is not to say that we can reject the presence of
endogeneity in the relationship between aid and alliances; rather it is
still mostly a matter of speculation rather than evidence one way or the
other.

Taking the above into consideration, while some variation in
donor-recipient ties is exogenous, some of its variation is presumably
endogenous to donor aid giving and, by extension, lead donorship. The
usual method of accounting for endogeneity is to take an instrumental
variables approach, however a convincingly exogenous instrument that
explains variation in trade and alliances that does not also explain
variation in lead donorship is hard to find. An alternative strategy is
simply to use a lagged version of the explanatory variable suspected of
endogeneity. In their study on the impact of aid on trade,
@martinez2017aid use a two-year lag of aid to account for endogeneity.
So as an additional robustness check, the main analysis was repeated
using a two-year lagged version of donor-recipient tie concentration in
place of the within-year version of the measure. The results are
visualized in Figure 11. As can be seen, the results remain unchanged.
While this does not mean we can rule out endogeneity as a major threat
to inference in this study, this finding is at least reassuring.

## What About Odd Cases of Lead Donorship?

The theory posits that concentrated marginal returns (donor-recipient
ties) are a necessary but insufficient condition for lead donorship.
Recipient need likewise is a necessary but insufficient condition for
lead donorship; with the caveat that donor behavior (cooperation or
cooperation failure) should determine whether recipient need is
proportional or inversely proportion to lead donorship under the
condition of concentrated marginal returns. The theory, and by extension
the hypotheses, are essentially Boolean in this regard because they are
dealing with a logical AND condition under which lead donorship is
predicted to arise with greatest frequency. However, it is important to
remember that this Boolean logic is not deterministic; it is
probabilistic. It therefore should not be a surprise to observe lead
donorship where the theory postulates that it should have a lower
probability of arising. Beyond this, the measurement strategy adopted in
this paper is not perfect. It is possible that it has blind spots, so
looking at exceptional cases can be instructive about where and why the
measurement strategy fails.

Figure 12 breaks the data down by ties concentration and need quartiles
and summarizes the frequency of lead donorship strength within each. The
counts shown indicate the number of recipient-years in the data where a
certain level of lead donorship is observed. It should come as no shock
that the frequencies are consistent with the results in the main
statistical analysis. Non-zero levels of lead donorship are observed
with the greatest frequency in the upper-right quadrant while cases with
no lead donor are most commonly observed when need and/or ties are low.
However, in no quartile does even the maximum level of lead donorship go
unobserved. Clearly, while the general pattern in the data is consistent
with the elusive collaboration hypothesis, there are clearly exceptional
cases.

```{r fig.height=6, fig.width=6, fig.cap="Frequency of lead donorship per recipient-year by ties concentration and need quartiles."}
agg_dt |>
  mutate(
    "Ties Concentration" = cut(
      ties_ssp,
      breaks = quantile(ties_ssp),
      labels = paste0(
        "Ties: ", c("1st", "2nd", "3rd", "4th")
      ),
      include.lowest = T,
      ordered_result = T
    ) |>
      factor(levels = paste0(
        "Ties: ", c("4th", "3rd", "2nd", "1st")
      )),
    "Need" = cut(
      need,
      breaks = quantile(need),
      labels = paste0(
        "Need: ", c("1st", "2nd", "3rd", "4th")
      ),
      include.lowest = T,
      ordered_result = T
    )
  ) |>
  group_by(
    `Ties Concentration`,
    `Need`
  ) |>
  count(lead_donor) |>
  ggplot() +
  aes(
    x = lead_donor,
    y = n
  ) +
  geom_col(
    fill = "gray",
    color = "black"
  ) +
  geom_point() +
  geom_line() +
  geom_text(
    aes(label = n),
    vjust = -0.3,
    fontface = "bold"
  ) +
  labs(
    x = "Strength of Lead Donorship",
    y = "Count"
  ) +
  scale_x_continuous(
    n.breaks = 4,
    labels = c("0", "1/3", "2/3", "1")
  ) +
  ylim(c(0, 165)) +
  facet_grid(
    `Ties Concentration` ~
      `Need`
  ) 
```

Consider cases in the fourth quartile of need and the first quartile of
ties concentration. The elusive collaboration hypothesis holds that
recipients in this subset of the data are subject to underfunding of aid
by donors in equilibrium, but that no one donor will be left to bear the
lion's share of the responsibility for doing so. However, there are
nonetheless recipient-years in this subset of the data with a positive
level of lead donorship. There are even 12 recipient-year observations
where lead donorship is at the maximum level of 1. A closer look at the
data reveals that this maximum lead donorship is observed in five unique
aid recipients: Cameroon, Haiti, Iraq, Sudan, and Vietnam. In three of
these donors, a lead donorship score of 1 arises in only one year for
each (1999 in Cameroon, 2006 in Iraq, and 1995 in Vietnam). Maximum lead
donorship is observed in Haiti in three years (2004, 2007, and 2009),
and it is observed in Sudan in six years (2004, 2005, 2007, 2008, 2009,
2011).

The identities of the lead donor in each of these cases is not
surprising. The lead donor in Cameroon was France in 1999. In fact,
France was the lead donor here in 1995 and 2000 as well, but these
observations for Cameroon fall outside of quartiles for need and ties
under consideration here. The reason is that need in Cameroon spikes in
1999 but is lower in 1995 and 2000 because of the 1999 eruption of Mt.
Cameroon that lead to disasterous earthquakes, many deaths, and
substantial internal displacement. Why is France the donor that has been
left to bear the greatest share of the burden in responding? This first
case reveals a clear limitation with the measurement strategy in this
study. Cameroon is a former colony of France, making the recipient of
clear importance and salience to French policymakers. Clearly, the
measure of donor-recipient ties does not adequately weight former
colonial status in the case of Cameroon.

Haiti, Iraq, and Sudan are also interesting cases that reveal possible
limitations with this study's measurement strategy. The United States is
Lead Donor in all of these countries (and for years beyond those
highlighted above). Haiti is a small country but one that has been
subject to notoriously systemic political instability, poverty, and
destruction from a combination of hurricanes and earthquakes. Even
though it is a former colony of France, it's proximity to the United
States means that the US is especially sensitive to negative
externalities from Haiti's tumultuous past and present. Clearly, this is
another case where the measure of donor-recipient ties fails to
appropriately weight one of the dimensions of these ties (distance).
Furthermore, other dimensions of ties such as trade may not be
substantial enough to help differentiate the US relationship with Haiti
compared to other donors since, as a poor country, total trade volume
between Haiti and other countries has a lower overall baseline compared
to trade relations between donors and other aid recipients.

Iraq and Sudan are exceptions for a different reason—their strategic
military importance to the US. This is a dimension of donor-recipient
ties that alliances is meant to capture, but this approach has clear
limitations. Countries can be the object military importance to donors
without the precence of a formal military alliance. Israel is an
important strategic partner with the US, and yet the two countries have
not formalized this relationship with a military pact. This blind spot
in the measure of recipient-donor ties reflects the unfortunate reality
that measures of this specific dimension of donor ties to recipients are
not readily available for most donor countries. The best data is for the
United States, both with respect to factors like troop deployments and
military aid. Such factors may be especially relevant for the United
States because of its unique role in global politics.

Finally, Japan is the lead donor in the case of Vietnam. This example
again reveals how distance as a dimension of donor-recipient ties may be
under-weighted in the measurement strategy. Compared to all the other
donor countries in the data, Japan has the closest proximity to Vietnam,
making it especially sensitive to needs that arise there and, thus, the
most likely to take the lead in responding to these needs while other
donors sit on the sidelines.

Taken together, this handful of examples shows that the measurement
strategy for identifying donor-recipient ties has blind spots. However,
in each example, these exceptional cases would seem to support the main
theoretical argument rather than conflict with it. That is, these are
exceptions that prove the rule. Clearly, donor-recipient ties vary in
ways that the measure constructed in this study is too blunt an
instrument to capture. This, admittedly, is a limitation with any method
that collapses multiple factors into one. This may be an important issue
for future research to address.

```{r eval=FALSE, include=FALSE}
agg_dt |>
  mutate(
    "Ties Concentration" = cut(
      ties_ssp,
      breaks = quantile(ties_ssp),
      labels = paste0(
        "Ties: ", c("1st", "2nd", "3rd", "4th")
      ),
      include.lowest = T,
      ordered_result = T
    ) |>
      factor(levels = paste0(
        "Ties: ", c("4th", "3rd", "2nd", "1st")
      )),
    "Need" = cut(
      need,
      breaks = quantile(need),
      labels = paste0(
        "Need: ", c("1st", "2nd", "3rd", "4th")
      ),
      include.lowest = T,
      ordered_result = T
    )
  ) |>
  filter(lead_donor == 1,
         `Need` == "Need: 4th",
         `Ties Concentration` == "Ties: 1st") |>
  group_by(recipient) |>
  summarize(
    n = n(),
    years = paste0(year, collapse = ", ")
  ) |> view()

dt |>
  filter(lead_donor == 1 &
         recipient %in% c("Haiti", "Sudan", "Iraq", "Viet Nam", "Cameroon")) |>
  view()
```

# Conclusion

This study began by asking a simple question: does lead donorship arise
because donor governments cooperate or fail to cooperate in the
allocation of foreign aid? This is a question that past research as
tried to address, but using spatial models and some strong assumptions
about how to interpret correlations in donor giving. While helpful, this
approach is subject to circular causation and unmeasured endogeneity and
heterogeneity in the spatial correlation in donor giving, raising
questions about how best to interpret the results. This study adopts a
different approach, modeling variation in lead donorship as the outcome
of interest using novel measures to triangulate the goals of donor
governments and, thus, to determine whether the distribution of lead
donorship in developing countries is consistent with donor cooperation
or systematic cooperation failure.

Following arguments laid out by @steinwand2015, if donors cooperate, we
should observe that they (1) agree to a delineation of spheres of
influence when and where non-development or rival goods dominate and (2)
forego free-riding when development or public goods dominate.
Alternatively, if donors fail to cooperate, they will (1) compete for
top spot when non-development or rival goods dominate and (2) engage in
free-riding when development or public goods dominate. The first pattern
is formalized in this study as the "collaboration" hypothesis and the
second as the "elusive collaboration" hypothesis.

The analysis presented here supports the second hypothesis. That is, the
empirical evidence consistent with a systematic absence of cooperation
among donors in the allocation of aid. Buttressing this conclusion are a
few important assumptions and new aggregate measures capturing variation
in donor-recipient ties and recipient development need that help in
triangulating when public goods are predominant and marginal returns
from either public or private goods are concentrated in a single donor.
While these measures are admittedly blunt instruments, what is most
important is that they can capture the right signals in the data. On
this front, these measures appear to succeed; albeit, with some
exceptions that show there is yet room for improvement.

The findings of this study are important for ongoing policy debates and
raise significant theoretical questions. Lead donorship remains
under-appreciated in the political economy literature, but as
@steinwand2015 argues the concept has important welfare implications,
both for donor countries and aid recipient countries. The lack of
cooperative behavior among donors on a macro level is frustrating,
considering the efforts made over decades to foster greater donor
cooperation in high-level fora. This study calls for further inquiry
into why collaboration remains elusive. Are there specific
characteristics of the strategic environment inhabited by donors and
recipients that hinder cooperation? Future work should also consider
whether the emergence of lead donorship really enhances or hampers aid
effectiveness.

Future research should also expand the scope of this analysis by
considering how the entrance of many so-called un-traditional donors
influences patterns in lead donorship. China in particular has emerged
as an important actor in international development over the past decades
[@dreher2022banking], and many other countries in South America, the
Middle East, and Asia have begun giving foreign aid as well
[@dreherEtAl2011]. Many of these emerging donors appear to behave
similarly to traditional donors, but as research by @dreherEtAl2011
shows, these donors seem to be less concerned about recipient need. This
suggests these donors may make donor cooperation in delineating spheres
of influence even more difficult. This is likely especially the case
with China since many policymakers in Western donors, particularly in
the United States, explicitly see China's foreign aid as a strategic
threat.[^3]

[^3]: See, for example, the policy recommendations of Andrew
    @natsios2020, former head of the United States Agency for
    International Development, and also the 2021 "Countering Chinese
    Communist Party Malign Influence Act" which called for raising funds
    in response to Chinese overseas development project (link:
    <https://www.congress.gov/bill/117th-congress/house-bill/2329/text?r=82&s=1>).

In sum, the existence of lead donorship in the allocation of foreign aid
does not appear to be good news about donor cooperation. However, this
finding will surely not be the final word on this issue. It is this
author's hope that future scholarship will refine and expand on the
analysis presented here and will further consider whether lead
donorship, since it appears to be the product of uncooperative behavior,
is bad for aid effectiveness.

\clearpage

# References

```{=tex}
\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\vspace*{-0.2in}
```
\noindent
